### 2016-07-24 paper/paper.tex
Yes I know im not supposed to write the abstract first

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\begin{document}
\begin{abstract}
In this paper, we describe a DSL for web APIs named Servis. Servis is heavily inspired by Servant, a haskell type-level DSL for web APIs,  but tries to solve identified shortcomings by using a dependently typed language called Idris. Servis uses the universe pattern to deduce types from an API description such that we can verify at compile time that the implementation of the API matches the description.

Furthermore, we take a look at if dependent types can bring more to the table in terms of expressiveness of API descriptions than the original Servant DSL. Specifically, we introduce $\pi$-types to describe interdependencies within an API description.

\end{abstract}
\end{document}

### 2016-07-31 paper/paper.tex
Work on paper

@@
 \usepackage{graphicx}
+\usepackage{minted}
 \usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
+\author{Arian van Putten \\ a.vanputten@uu.nl}
+\title{Servis: A dependently typed DSL for web APIs}
 \begin{document}
+\maketitle
 \begin{abstract}
@@
 
-Furthermore, we take a look at if dependent types can bring more to the table in terms of expressiveness of API descriptions than the original Servant DSL.
+Furthermore, we take a look at if dependent types can bring more to the table in terms of expressiveness of API descriptions than the original Servant DSL\@.
 Specifically, we introduce $\pi$-types to describe interdependencies within an API description.
 
+  TODO: rewrite abstract \emph{after} I finish the paper :-)
+
 \end{abstract}
+
+\section{Introduction}
+Servant is an extensible type-level DSL for Haskell, that tries to make web APIs \emph{first-class}~\cite{servant}.
+A web API should be a thing that can be named, passed to, or returned by functions.
+Just like any other term in a language.
+
+It does this by describing the DSL at the type-level.
+And then uses \emph{type families} (type-level functions) to be able to perform computations on the API specifications.
+Using these type families, we are able to write different \emph{interpretations} of an API\@.
+Example of interpretations are: generating documentation from a specification, generating types of a server implementation, or even generating a client that can interfact with the API\@.
+
+In this paper, I will describe a port of Servant in Idris, and explore what kind of benefits this brings us.
+
+\subsection{An introduction to Servant}
+I will now give a very short overview of the internals of Servant. We will start by describing the type-level DSL of servant.
+
+\begin{listing}[H]
+\begin{minted}{haskell}
+data api1 :<|> api2
+infixr 8 :<|>
+
+data (item :: k) :> api
+infixr 9 :>
+
+data ReqBody (ctypes :: [*] (t :: *)
+data Capture (symbol :: Symbol) (t :: *)
+
+data Get (ctypes :: [*]) (t :: *)
+data Post (ctypes :: [*]) (t :: *)
+data JSON
+\end{minted}
+\caption{The Servant DSL}
+\label{lab:servant}
+\end{listing}
+
+TODO give a short overview of servant
+
+Servant has a few limitations which we will describe in Section~\ref{sec:limitations}.
+
+\subsection{Why idris}
+Idris is a language that in syntax is rather similar to Haskell, but has a more powerful type system.
+Idris is a dependently-typed language.
+This means that types are fist-class citizens.
+We can pass types to functions, let functions create new types, and let types depend on other value-level terms.
+
+Though haskell has type-level functions (called type-families), it does not allow these functions to dpend on term-level values.
+There is a distinct boundary between type-level and term-level, whilst the two are merged in Idris.
+This makes doing computations with types a lot more natural (no special constructs are needed, just write functions), but also lets us do some more powerful stuff like letting types actually depend on runtime user input.
+
+Dependent Haskell is currently in the making, and probably planned for around GHC version 8.6~\cite{neeeded}.
+In the future, it would probably be very feasible to port this library back to Haskell and bring all the benefits along with it.
+
+\subsection{Contributions}
+In this paper, I make the following contributions:
+
+TODO reword
+\begin{itemize}
+  \item I point out several problems an limitations of the original Servant DSL
+  \item I show a port of servant to Idris, a dependently-typed language
+  \item I show how new techniques from Idris can help mitigate some of the problems that Servant has.
+\end{itemize}
+
+\section{Limitations of Servant}\label{sec:issues}
+\subsection{Term and type-level are not unified in Haskell}
+As explained in the introduction, the Servant DSL lives in the type-level.
+In haskell, types and terms are separated.
+One can lift terms into the type-level using GHC extensions like \texttt{XDataKinds}, but that only gets us so far.
+On the term level, Haskell has many great libraries and data structures that can be used.
+For example, all kinds of string manipulations.
+It would be useful to be able to use these functions to manipulate APIs, making them truly first-class.
+Now they are only first-class in the type level.
+
+For example, a useful function would be \mintinline{haskell}{path "/users/friends/etc" = "users" :> "friends" :> "etc"} to reduce noise in the DSL\@.
+But this is currently impossible to write because Strings in the type level are of kind \mintinline{haskell}{Symbol} and not \mintinline{haskell}{String} so all our favorite string manipulation functions like \mintinline{haskell}{splitOn "/"} will not work.
+This would mean that we would have to re-implement all kinds of utility functions on the type-level.
+
+\subsection{A kind of APIs}
+As mentioned in Section~\ref{sec:servant_intro}, a delibirate choice has been made to let the Servant DSL live in the open kind \mintinline{haskell}{*}.
+Though this makes the DSL extensible, it also makes the DSL fragile and prone to error.
+
+To give a real-world example, here is a snippet from a user on the Haskell Subreddit that could not get his API to work:
+
+\begin{listing}[H]
+\begin{minted}{haskell}
+type SoundcloudTrackAPI = "tracks" :> 
+  (    QueryParam "client_id" T.Text 
+    :> QueryParams "genres" T.Text
+    -> Get [JSON] [ST.Track] 
+  :<|> QueryParam "client_id" T.Text
+    :> Capture "id" Int
+    :> Get [JSON] ST.Track
+  )
+\end{minted}
+\caption{A malformed API}
+\label{code:problem}
+\end{listing}
+
+The problem is hard to spot, but the user typed wrote \mintinline{haskell}{->} instead of \mintinline{haskell}{:>}.
+This code compiles fine though.
+Because the Servant DSL lives in kind \mintinline{haskell}{*}, \mintinline{haskell}{:>} is   of kind \mintinline{haskell}{* -> * -> *}.
+Which is the same kind as the kind of \mintinline{haskell}{->}.
+
+Only when one would try to interpret the DSL, like in Listing~\ref{code:interpret}, one would get a type error because no type family pattern matches on \mintinline{haskell}{->}.
+
+\begin{listing}
+\begin{minted}{haskell}
+searchTracksByGenre :: Maybe T.Text 
+                    -> [T.Text]
+                    ->  Manager
+                    -> BaseUrl
+                    -> ClientM [ST.Track]
+getTrack            :: Maybe T.Text
+                    -> Int
+                    -> Manager
+                    -> BaseUrl
+                    -> ClientM ST.Track
+(searchTracksByGenre :<|> getTrack) = client soundcloudAPI 
+\end{minted}
+  \caption{An interpretation of the API (Listing ~\ref{code:problem})}
+  \label{code:interpret}.
+\end{listing}
+
+Preferably, one would want to define a new open kind \mintinline{haskell}{API} such that \mintinline{haskell}{(:>) :: API -> API -> API}, but this is not possible in Haskell (or Idris).
+Instead, one could define a closed kind (using \textt{XDataKinds}), in which we define the entire Servant DSL.
+This would mean we would lose extensibility of the DSL, but we gain kind-safety.
+
+Also, as described in Section~\ref{sec:servant_intro}, the Servant DSL is split up in several semantic portions.
+For example, \mintinline{haskell}{:>} should take an \mintinline{haskell}{item} on the left-hand side, which describes either a constraint like \mintinline{haskell}{ReqBody} or a path-piece, whilst the right-hand side should be an \mintinline{haskell}{api}.
+These semantic portions are implict and not checked though.
+So we can create totally non-sensical API types like \mintinline{haskell}{type Nonsense = (Int :>Int) :<|> ("hey ":> Int)} that have no interpretation at all.
+
+Instead of having these implicit semantic rules about what kind of arguments an operator can take in the Servant DSL, we could define a distinct closed kind for each semantic portion.
+Such that the type of \mintinline{haskell}{:>} changes to the following: \mintinline{haskell}{(:>) :: Item -> API -> API}.
+
+Another implict rule is, that a chain of \mintinline{haskell}{:>} should always end in a \mintinline{haskell}{Get} or a \mintinline{haskell}{Post}.
+So this should be invalid: \mintinline{haskell}{type API = "a" :> ReqBody [JSON] User}.
+This can also not be checked if we do not add more fine-grained kinds to the DSL.
+
+\subsection{Explicit status codes}
+An API that describes how stuff goes right is good, and API that describes how stuff goes wrong is great.
+But currently, Servant has no way to indicate in the API description what kind of status codes an endpoint can return.
+Thus this also does not show up in the generated documentation.
+
+It would be nice to add a list of possible status codes that an endpoint can returns, and then also be able to check whether a server implementation actually only returns one of the specified status codes if something goes wrong (or right).
+
+\subsection{Overlapping routes}
+Another problem with Servant is that there is no protection for overlapping routes. If two routes could resolve to the same path, Servant makes a left-biased choice~\cite{needed}.
+It would be nice if we could verify at compile time that no routes in an API specification can overlap.
+
+\begin{listing}
+\begin{minted}{haskell}
+  type API  = "users" :> Capture "args" String :> Get [JSON] User
+         :<|> "users" :> "favorites" :> GET [JSON] User
+\end{minted}{haskell}
+\caption{An example of overlapping routes. The capture will be chosen if args equals "favorites"}
+\end{listing}
+
+\section{Servis}
+In this section, I present a DSL in Idris named Servis.
+It tries to address each issue listed in Section~\ref{sec:issues}.
+Firstly, we will define the DSL in which we can write API specifications.
+Then, we will look at how to write an interpreter of this DSL that generates documentation.
+Then, we will look at how we would implement HTTP webserver handlers that adhere to the DSL.
+Next, we'll look at solving the overlapping routes problem by writing proofs in Idris.
+Finally, we will extend the DSL to add \emph{pi-types}, allowing us to describe dependencies in our API\@.
+For example, one might want to describe  the size of a list of users in the API, but let the size depend on a query parameter \emph{limit} that limits the response size.
+It would then force the implementor of the API to only return values smaller or equal to limit in size.
+
+\subsection{The DSL}
+\subsection{A simple interpreter: Documentation generation}
+\subsection{An interpreter for an HTTP Server}
+\subsubsection{Checking explicit status codes}
+\subsection{Pi-types}
+\subsection{New problems that pi-type create}
+
+\section{Summary}
+\section{Related work}
+Talk about tel/serv. An inspiration for this project.
+\section{Conclusions and future work}
+
 \end{document}

### 2016-07-31 paper/paper.tex
More paperwork

@@
 Then, we will look at how to write an interpreter of this DSL that generates documentation.
-Then, we will look at how we would implement HTTP webserver handlers that adhere to the DSL.
+Then, we will look at how we would implement HTTP webserver handlers that adhere to specification defined by the API DSL\@.
 Next, we'll look at solving the overlapping routes problem by writing proofs in Idris.
@@
 \subsection{The DSL}
+Here we will, step by step, describe the DSL in which we describe APIs.
+Furthermore, we will address the problem of the open-kindedness of Servant.
 \subsection{A simple interpreter: Documentation generation}
+Here we will show a simple interpretation of the DSL, by generating some basic documentation.
+It's basically a function \mintinline{idris}{docs : API -> String}. 
 \subsection{An interpreter for an HTTP Server}
+A more advanced example where we define two interpreters.
+One 'interpreter' that generates types from our DSL such that we can type-check our server handler implementations.
+(The universe pattern), and another interpreter that routes requests to these handler implementations.
 \subsubsection{Checking explicit status codes}
+Shows that we can enforce explicit status codes
+
+\subsection{A proof of non-overlapping routes}
+We will describe types for helping prove that routes in an API definition don't overlap.  We do this
+by providing Decidable Equality proofs for all parts of our DSL.
+\begin{minted}{idris}
+
+||| A Proof that the GET dtor is injective over its requestType
+injGET : (GET x = GET y) -> x = y
+injGET Refl = Refl
+
+||| A proof that the POST dtor is injective over its requestType
+injPOSTReq : (POST requestType1 responseType1 = POST requestType2 responseType2) -> requestType1 = requestType2
+injPOSTReq Refl = Refl
+
+||| A Proof that the POST dtor is injective over its responseType
+injPOSTResp : (POST requestType1 responseType1 = POST requestType2 responseType2) -> responseType1 = responseType2
+injPOSTResp Refl = Refl
+
+||| A Proof that a GET is not a POST
+getNotPost : (GET responseType = POST requestType responseType1) -> Void
+getNotPost Refl impossible
+
+||| A proof that GET is congruent
+cong_GET : (prf : responseType1 = responseType2) -> GET responseType1 = GET responseType2
+cong_GET = cong
+
+||| A proof POST is congruent
+cong_POST : (prf2 : responseType1 = responseType2) -> (prf1 : requestType1 = requestType2) -> POST requestType1 responseType1 = POST requestType2 responseType2
+cong_POST Refl Refl = Refl
+
+(DecEq req, DecEq res) => DecEq (Handler req res) where
+  decEq (GET responseType1) (GET responseType2) =
+    (case decEq responseType1 responseType2 of
+          (Yes prf) => Yes (cong prf)
+          (No contra) => No ((\h => contra (injGET h))))
+  decEq (POST requestType1 responseType1) (POST requestType2 responseType2) =
+    (case decEq requestType1 requestType2 of
+          (Yes prf1) => (case decEq responseType1 responseType2 of
+                             (Yes prf2) => Yes (cong_POST prf2 prf1)
+                             (No contra) => No (\h => contra (injPOSTResp h)))
+          (No contra) => No (\h => contra (injPOSTReq h)))
+  decEq (GET _) (POST _ _) = No getNotPost
+  decEq (POST _ _) (GET _) = No (negEqSym getNotPost)
+data DisjointPP : PathPart capture query -> PathPart capture query -> Type where
+  ConstD : Not (str = str') -> DisjointPP (Const str) (Const str')
+
+data DisjointPath : Path capture query req res -> Path capture query req res -> Type where
+  OutputsD : Not (handler1 = handler2) -> DisjointPath (Outputs handler1) (Outputs handler2)
+  PathBase : DisjointPP pp1 pp2 -> DisjointPath (pp1 :> p1) (pp2 :> p2)
+  PathStep : DisjointPath p1 p2 -> DisjointPath (p :> p1) (p :> p2)
+
+data DisjointAPI : List (Path capture query req res) -> Type where
+  Base : DisjointAPI []
+  Step : (x : Path capture query req res) -> All (DisjointPath x) xs -> DisjointAPI (x::xs)
+\end{minted}
 \subsection{Pi-types}
+Adds dependent pairs to our DSL such that we can describe dependencies between query params, captures and return types.  Example:
+
+\begin{minted}{haskell}
+exampleOfDependentPath : Path EmptyU QueryU EmptyU RespU
+exampleOfDependentPath =
+  QueryParam "limit" NAT :*> \limit => Outputs (GET (BLIST limit USER))
+
+getUsersLimit : (limit : Nat) -> IO (BoundedList limit User)
+getUsersLimit limit = pure . take limit $ users
+
+exampleOfDependentPathHandler : el Example.exampleOfDependentPath
+exampleOfDependentPathHandler = getUsersLimit
+
+\end{minted}
 \subsection{New problems that pi-type create}
+Here we will describe how pi-types adds problems. For example we cannot easily generate documentation anymore, prove that there are no overlapping routes, or prove that statuscodes are adhered to.
 

### 2016-08-21 paper/paper.tex
Work on text

@@
 \usepackage{graphicx}
+\usepackage{hyperref}
 \usepackage{minted}
@@
 
+\subsection{The Universe pattern}
+The Servis DSL makes use of the Universe pattern to map an API description to types of handlers.
+ A universe is a type \mintinline{idris}{U : Type} which contains names for types and a type family \mintinline{idris}{el : U -> Type} that assigns to every name \mintinline{idris}{a : U} the type of its elements \mintinline{idris}{el a : Type} \cite{generic}\@.
+
+ In idris, we could define this as the following interface:
+ \begin{listing}
+ \begin{minted}{idris}
+interface Universe (u : Type) where
+  el : u -> Type
+ \end{minted}{idris}
+   \caption{Universe interface in Idris}
+ \end{listing}
+
+ In our case, we have a datatype \mintinline{idris}{API : Type}, which describes our API, and \mintinline{idris}{el} maps values to handler types. A Short example:
+ \begin{minted}{idris}
+
+-- we define a value of API
+api : API
+api = Const "users" :> Capture Int "userId" :> Outputs (Get USER)
+
+getUser : Int -> IO User
+getUser = ?someDatabaseMagic
+
+--  el api evaluates to  Int -> IO User
+handler : el api
+handler = getUser
+ \end{minted}
 \subsection{The DSL}
-Here we will, step by step, describe the DSL in which we describe APIs.
-Furthermore, we will address the problem of the open-kindedness of Servant.
+So what does this type \mintinline{idris}{API} look like?
+\subsubsection{Handler}
+Before we will look at routes, lets think of what an endpoint should look like.
+We should be able to handle the HTTP verbs. We will only implement \mintinline{idris}{GET} and \mintinline{idris}{POST}, because the others are very similar.
+For a \mintinline{idris}{GET}, we should be able to specify a return type of the handler,  and for a \mintinline{idris}{POST}, we should be able to
+specify a request body type, and a return type.
+
+A first try at describing this as a datatype for our universe  could look something like this:
+\begin{listing}
+  \begin{minted}{idris}
+data Handler : Type where
+  GET : (responseType : Type) -> Handler req res
+  POST : (requestType : Type) -> (responseType : Type) -> Handler req res
+
+Universe Handler where
+  -- a GET handler just returns something of type responseType
+  el (GET responseType) = IO responseType
+  -- a POST handler takes the requestBody as argument, and returns a response
+  el (POST requestType responseType) = requestType -> IO responseType
+
+  \end{minted}
+\end{listing}
+
+We could then define a handler for \mintinline{idris}{GET}ting users as follows:
+\begin{minted}{idris}
+record User where
+  constructor MkUser
+  name : String
+  email : String
+
+userHandler : Handler
+userHandler = GET User
+\end{minted}
+
+However, once we start implementating an actual server in Section~\ref{sec:server}, we will discover that this definition is not ideal. What if we want to write some
+function that does something different based on the \mintinline{idris}{responseType} of a \mintinline{idris}{GET}?
+For example, based on value of \mintinline{idris}{responseType}, we need to decide how to encode the response body to JSON.
+Lets imagine a hypothetical function \mintinline{idris}{selectEncoder} that tries to implement this behaviour:
+
+\begin{minted}{idris}
+selectEncoder : Type -> Encoder
+selectEncoder User = ?jsonEncoderForAUser
+
+userEncoder : Encoder
+userEncoder = selectEncoder User
+\end{minted}{idris}
+
+Sadly enough, this will give us a compiler error. The problem is that in Idirs, we cannot pattern match on values of \mintinline{idris}{Type}. 
+However, this exact problem can be solved with universes! We can define a data type \mintinline{idris}{ResponseUniverse : Type} which names
+the \mintinline{idris}{User} type. We can then pattern-match on \mintinline{idris}{ResponseUniverse} instead of \mintinline{idris}{Type}!
+And whenever we need the \mintinline{idris}{User} type, we can just call \mintinline{idris}{el USER : Type}.
+
+\begin{minted}{idris}
+data ResponseUniverse : Type where
+  USER : ResponseUniverse
+
+Universe ResponseUniverse where
+  el USER = User
+
+selectEncoder : ResponseUniverse -> Encoder
+selectEncoder USER = ?jsonEncoderForAUser
+
+userEncoder : Encoder
+userEncoder : selectEncoder USER
+\end{minted}
+
+This leads us to the following definition of \mintinline{idris}{Handler}:
+
+\begin{listing}
+  \begin{minted}{idris}
+data Handler : (req : Type)  -> (res : Type) -> Type where
+  GET : (responseType : res) -> Handler req res
+  POST : (requestType : req) -> (responseType : res) -> Handler req res
+
+(Universe req, Universe res) => Universe (Handler req res) where
+  el (GET responseType) = IO (el responseType)
+  el (POST requestType responseType) =
+    el requestType -> IO (el responseType)
+  \end{minted}
+  \caption{The new Handler definition}
+\end{listing}
+
+\subsubsection{Path and PathPart}
+Of course we do not only get request information from the request body,
+we also get information about a request by parsing it's request URL. This
+is where \mintinline{idris}{Path}s and \mintinline{idris}{PathPart}s come into play. 
+When we talk about path parts, we mean sections of an path template that convey parsable information.
+For example, the following path template has three path parts: \texttt{/users/\string{user\_id\string}?search=\string{query\string}}. A constant piece
+in the path \texttt{users}, a variable capture \texttt{user\_id} and a query param \texttt{search}. This
+decomposition of a path template can be written down as a datatype. In this datatype we do not only
+want to describe the names of the parts of these paths, but also their types. Hence, we will parameterize
+over a universe similarly to how we defined \mintinline{idris}{Handler}.
+\begin{minted}{idris}
+
+||| PathPart is a part of a path.
+||| @ capture the universe of captures
+||| @ query   the universe of query params
+data PathPart : (capture : Type) -> (query : Type) -> Type where
+  ||| A constant piece of text
+  ||| @ path the path part
+  Const : (path : String) -> PathPart capture query
+  ||| A capture of a variable name of a specific type
+  ||| @ name  the name of the variable
+  ||| @ type  the type of the variable
+  Capture : (name : String) -> (type : capture) -> PathPart capture query
+  ||| A query param of a variable name of a specific type
+  ||| @ name  the name of the variable
+  ||| @ type  the type of the variable
+  QueryParam : (name : String) -> (type : query) -> PathPart capture query
+\end{minted}
+
+A \mintinline{idris}{Path} is simply a list if \mintinline{idris}{PathPart}s, which ends in a \mintinline{idris}{Handler}:
+\begin{minted}{idris}
+||| Describes a Path. A Path consists of path parts followed by a handler
+data Path : (capture : Type) -> 
+            (query : Type) ->
+            (req : Type) ->
+            (res : Type) -> Type where
+  ||| Ends a path with a handler
+  ||| @ handler the handler
+  Outputs : (handler : Handler req res) -> Path capture query req res
+  ||| Conses a pathpart to a path
+  (:>) : PathPart capture query -> Path capture query req res -> Path capture query req res
+infixr 5 :>
+\end{minted}
+
+The Path we just described can now be written as
+\begin{minted}{idris}
+   Const "users"
+:> Capture "user_id" INT
+:> QueryParam "search" STRING
+:> Outputs (GET USER)
+\end{minted}
+
+The universe implementation is straightforward for paths, except for
+that we do not wannt to have spurious arguments in case when we encounter
+a \mintinline{idris}{Const}:
+\begin{minted}{idris}
+( Universe capture
+, Universe query
+) => Universe (PathPart capture query) where
+  el (Const path) = ()
+  el (Capture name type) = el type
+  el (QueryParam name type) = el type
+
+( Universe capture
+, Universe query
+, Universe req
+, Universe resp
+) => Universe (Path capture query req resp) where
+  -- special case because we don't like () in our handler functions
+  el (Const path :> right) =  el right
+  el (pathPart :> right) = el pathPart -> el right
+  el (Outputs handler) = el handler
+\end{minted}
+
+\subsubsection{The API Datatype}
+Finally, we define an \mintinline{idris}{API} as an non-empty list of \mintinline{idris}{Path}s.
+
+\begin{minted}{idris}
+data API : (capture : Type) -> (query : Type) -> (req : Type) -> (res : Type) -> Type where
+  OneOf : (paths : Vect (S n) (Path capture query req res)) -> API capture query req res
+\end{minted}
+
+The universe implementation of an API is the type of a list of handlers.
+Every handler is of a different type though, so instead of using an ordinary list,
+we use an heterogenous list, where each element \mintinline{idris}{x} in \mintinline{idris}{xs}
+is of type \mintinline{idris}{el x}.
+\begin{minted}
+( Universe capture
+, Universe query
+, Universe req
+, Universe res
+) => Universe (API capture query req res) where
+  el (OneOf xs) = HVect (map el xs)
+
 \subsection{A simple interpreter: Documentation generation}
@@
 It's basically a function \mintinline{idris}{docs : API -> String}. 
-\subsection{An interpreter for an HTTP Server}
+\subsection{An interpreter for an HTTP Server}\label{sec:server}
 A more advanced example where we define two interpreters.

### 2016-08-22 paper/paper.tex
work

@@
 \subsection{Overlapping routes}
-Another problem with Servant is that there is no protection for overlapping routes. If two routes could resolve to the same path, Servant makes a left-biased choice~\cite{needed}.
+Another problem with Servant is that there is no protection for overlapping routes. If two routes could resolve to the same path, Servant makes a left-biased choice~\cite{servant}.
 It would be nice if we could verify at compile time that no routes in an API specification can overlap.
@@
 The Servis DSL makes use of the Universe pattern to map an API description to types of handlers.
- A universe is a type \mintinline{idris}{U : Type} which contains names for types and a type family \mintinline{idris}{el : U -> Type} that assigns to every name \mintinline{idris}{a : U} the type of its elements \mintinline{idris}{el a : Type} \cite{generic}\@.
+ A universe is a type \mintinline{idris}{U : Type} which contains names for types and a type family \mintinline{idris}{el : U -> Type} that assigns to every name \mintinline{idris}{a : U} the type of its elements \mintinline{idris}{el a : Type} \cite{Altenkirch}\@.
 
@@
 api : API
-api = Const "users" :> Capture Int "userId" :> Outputs (Get USER)
+api = Const "users" :> Capture Int "userId" :> Outputs (GET User)
 
@@
 So what does this type \mintinline{idris}{API} look like?
-\subsubsection{Handler}
+\subsubsection{Handler}\label{sec:handler}
 Before we will look at routes, lets think of what an endpoint should look like.
@@
   ||| Conses a pathpart to a path
-  (:>) : PathPart capture query -> Path capture query req res -> Path capture query req res
+  (:>) : PathPart capture query ->
+         Path capture query req res ->
+         Path capture query req res
 infixr 5 :>
@@
 
-The universe implementation of an API is the type of a list of handlers.
+The universe implementation of an \mintinline{idris}{API} is the type of a list of handlers.
 Every handler is of a different type though, so instead of using an ordinary list,
@@
   el (OneOf xs) = HVect (map el xs)
+\end{minted}.
+
+\subsection{An example API}
+Using this DSL, we will now describe an API for interacting with users in a database.
+We will start of with defining our domain objects, and intercations we could do with them.
+
+\begin{listing}
+\begin{minted}{idris}
+record User where
+  constructor MkUser
+  id : Int
+  name : String
+  email : String
+  password : String
+
+findUserByName : String -> IO (List User)
+getUsers : (limit : Int) -> IO (List User)
+getUserById : (id : Int) -> IO User
+addUser : User -> IO ()
+\end{minted}
+\label{domain}
+\caption{The User object and its interactions}
+\end{listing}
+
+Next, we will describe the HTTP API for interacting with these users.  This includes
+the need of coming up with universes for captures, query params, request body, and response.
+
+\begin{listing}
+\begin{minted}{idris}
+data QueryUniverse
+  = QSTRING
+  | QINT
+Universe QueryUniverse where
+  el QSTRING = String
+  el QINT = Int
+
+data CaptureUniverse
+  = CINT
+Universe CaptureUniverse where
+  el CINT = Int
+
+-- Note: We use the same universe for requests and responses.
+data ReqResUniverse
+  = USER
+  | LIST ReqResUniverse
+Universe ReqResUniverse where
+  el USER = User
+  el (LIST a) = List (el a)
+
+userAPI : API QueryUniverse CaptureUniverse ReqResUniverse ReqResUniverse
+userAPI = OneOf
+  [ Const "users" :> Capture "user_id" CINT :> Outputs (GET USER)
+  , Const "users" :>
+    Const "find" :>
+    QueryParam "email" QSTRING :>
+    Outputs (GET (LIST USER))
+  , Const "users" :> Outputs (POST USER)
+  , Const "users" :>
+    QueryParam "limit" QINT :>
+    Outputs (GET (LIST USER))
+  ]
+\end{minted}
+\label{code:api}
+\caption{API definition for users}
+\end{listing}
 
 \subsection{A simple interpreter: Documentation generation}
-Here we will show a simple interpretation of the DSL, by generating some basic documentation.
-It's basically a function \mintinline{idris}{docs : API -> String}. 
+Writing an interpreter for generating documentation is very easy.
+Our API description isn't some type-level definition, it's just a plain old datatype that can be passed to functions and can be manipulated.
+
+We start by introducing an interface for docs:
+\begin{minted}{idris}
+interface HasDocs a where
+  docs :: a -> String
+\end{minted}
+Now, we just define an implementation for \mintinline{idris}{API} for this interface and we're done. Which is just calling \mintinline{idris}{docs} recursively on all its children.
+
+\begin{minted}{idris}
+(HasDocs req, HasDocs res) => HasDocs (Handler req res) where
+  docs (GET responseType) =
+    "## GET\nResponse type: " ++ docs responseType
+
+  docs (POST requestType responseType) =
+    "## POST\nRequest type: "  ++ docs requestType ++ "\n" ++
+    "Response type: " ++ docs responseType ++ "\n"
+
+renderParams : HasDocs query => List (String, query) -> String
+renderParams = renderParams' . reverse
+  where
+    renderParams' : List (String, query) -> String
+    renderParams' [] = ""
+    renderParams' ((a, b) :: []) =
+      "?" ++ a ++ "=<"++ docs b ++ ">"
+    renderParams' ((a, b) :: xs) =
+      renderParams' xs ++ "&" ++ a ++ "=<"++ docs b ++ ">"
+
+renderPath : 
+  ( HasDocs capture
+  , HasDocs query
+  , HasDocs req
+  , HasDocs resp) => Path capture query req resp -> String -> String
+renderPath (Outputs x) params = 
+    params ++ "\n" ++ docs x
+renderPath ((Const x) :> y) params =
+    "/" ++ x ++ renderPath y params
+renderPath ((Capture name type) :> y) params =
+  "/<" ++ name ++ ":" ++ docs type ++ ">" ++ renderPath y params
+renderPath _ _ = ""
+
+( HasDocs capture
+, HasDocs query
+, HasDocs req
+, HasDocs resp
+) => HasDocs (Path capture query req resp) where
+  docs path =
+    "# " ++ renderPath path (renderParams . getParams $ path) ++ "\n"
+    where
+
+      getParams : Path capture query req resp -> List (String, query)
+      getParams (Outputs x) = []
+      getParams ((Const x) :> y) = getParams y
+      getParams ((Capture name type) :> y) = getParams y
+      getParams ((QueryParam name type) :> y) = (name, type) :: getParams y
+
+( HasDocs capture
+, HasDocs query
+, HasDocs req
+, HasDocs resp
+) => HasDocs (API capture query req resp) where
+  docs (OneOf (x::[])) = docs x
+  docs (OneOf (x::y::xs)) = docs x ++ "\n" ++ oneOf (y::xs)
+\end{minted}
+
+Now \mintinline{idris}{docs userAPI}, assuming that the universes implement
+the \mintinline{idris}{HasDocs} interface, will evaluate to a very basic form of documentation:
+\begin{verbatim}
+# /users/<user_id:CINT>
+## GET
+ Response type: USER
+# /users/find?email=<QSTRING>
+## GET
+ Response type: LIST USER
+
+# /users?limit=<QINT>
+## GET
+ Response type: LIST USER
+
+# /users
+## POST
+Request type: USER
+Response type: USER
+
+# /users?limit=<QINT>
+## GET
+ Response type LIST USER
+\end{verbatim}
+
 \subsection{An interpreter for an HTTP Server}\label{sec:server}
-A more advanced example where we define two interpreters.
-One 'interpreter' that generates types from our DSL such that we can type-check our server handler implementations.
-(The universe pattern), and another interpreter that routes requests to these handler implementations.
-\subsubsection{Checking explicit status codes}
-Shows that we can enforce explicit status codes
+Now that we've shown that it is easy to manipulate APIs because they're just datatypes, it's time for a more complicated example.
+Though it is cool that we can generate documentation from an API description, the useful part that we want to implement is an HTTP Server.
+
+\subsubsection{A type for web applications}
+Lets first ask the question how a web server should look like in Idris. In the Servant paper, they describe the type of http handlers in the \texttt{wai} library, which looks like \cite{servant}:
+\begin{minted}{haskell}
+type Application = Request -> IO Response
+\end{minted}.
+
+We will more-or-less use this representation to model our HTTP handlers.
+they are  simply a function from request to response.
+The actual implementation of a web server based on this model is out of scope for now, but it would be interesting to port something like Haskells' \texttt{warp} (which impelements this concept) to Idris.
+
+Instead of having two abstract types \mintinline{idris}{Request} and \mintinline{idris}{Response}, we will assume for simplicity that a \mintinline{idris}{Response} is just a \mintinline{idris}{String}, and that a \mintinline{idris}{Request} is an \mintinline{idris}{URL}, and a requestBody of type \mintinline{idris}{Maybe String}.
+
+\begin{minted}{idris}
+Application : Type
+Application = (url : URL) -> (requestBody : Maybe String) -> IO String
+\end{minted}
+
+Just like in the Servant paper, we will need to modify this type a little bit more to suit our needs.
+We need to be able to express that an \mintinline{idris}{Application} does not match in case of a routing mismatch, such that we can try another route encompassed in \mintinline{idris}{OneOf}.
+
+Instead of returning  a value of \mintinline{idris}{IO String}, we can return a value of \mintinline{idris}{Maybe (IO String)}, where \mintinline{idris}{Just} signals a route match, whilst \mintinline{idris}{Nothing} signals a route mistmatch such that we can try another route.
+
+\begin{minted}{idris}
+RoutingApplication : Type
+RoutingApplication = (url : URL) ->
+                     (requestBody : Maybe String)
+                     -> Maybe (IO String)
+\end{minted}
+
+\subsubsection{Decoding and encoding data}
+In section~\ref{sec:handler}, we showed that if we want to implement some decoder, we would have to resort
+to the universe pattern. So that is what we will do exactly here.
+
+As an example, we will show how we would decode a queryparam. All other decoders follow the same pattern.
+What we want to do is, given a \mintinline{idris}{queryParam : String}, and a value \mintinline{idris}{v : u},
+where \mintinline{idris}{u} is the universe of query param types, and  \mintinline{idris}{v} a name of a type in that universe,
+perform the correct decoding to get a value of type \mintinline{idris}{el v}.
+
+We can summarize this behaviour in an interface:
+\begin{minted}{idris}
+interface Universe u => FromQueryParam u where
+  fromQueryParam : (v : u) -> String -> Maybe (el v)
+\end{minted}
+
+similarly, we have interfaces for captures and request bodies:
+\begin{minted}{idris}
+interface Universe u => FromCapture u where
+  fromCapture : (v : u) -> String -> Maybe (el v)
+
+interface Universe u => FromRequest u where
+  fromRequest : (v : u) -> String -> Maybe (el v)
+\end{minted}
+
+An implementation of \mintinline{idris}{FromQueryParam} for the universe \mintinline{QueryUniverse} in Listing~\ref{code:api}, would look like this:
+
+\begin{minted}{idris}
+FromQueryParam QueryU where
+  fromQueryParam QINT str = maybeParseInt str
+\end{minted}
+
+The fact that we can pattern match on the universe lets us inspect what type we need to return,
+and select the right decoder accordingly.
+
+We can use the same technique for encoding responses:
+\begin{minted}{idris}
+interface Universe u => ToResponse u where
+  toResponse : (v : u) -> el v -> String
+
+\end{minted}
+
+For example, we could use this to encode values in the \mintinline{idris}{ReqResUniverse} universe from
+Listing~\ref{code:api}:
+\begin{minted}{idris}
+  ToResponse ReqResUniverse where
+    toResponse USER (MkUser id name email password) =
+      "MkUser " ++ show id ++ " " ++ show name
+    toResponse (LIST elem) xs =
+      foldr (++) ""  . map (toResponse elem) $ xs
+\end{minted}
+
+\subsubsection{Routing requests}
+We will define the following interface to implement routing requests:
+\begin{minted}{haskell}
+interface Universe u => HasServer u where
+  route : (v : u) -> (handler : el v) -> RoutingApplication
+\end{minted}
+Or if we expaned the definition \mintinline{idris}{RoutingApplication}:
+\begin{minted}{haskell}
+interface Universe u => HasServer u where
+  route : (v : u) ->
+          (handler : el v) ->
+          (url : URL) ->
+          (requestBody : Maybe String)
+          -> Maybe (IO String)
+\end{minted}
+
+The idea of the router is, that given some part of the API \mintinline{idris}{v : u}, and a handler of type \mintinline{idris}{el v}, build an application that can either route succesfully or fail.
+Preferably, the application uses the handler parameter in its implementation of course.
+
+Lets start by defining an \mintinline{idris}{HasServer} instance for \mintinline{idris}{Handler}. Its task is to simply encode responses to Strings, and to handle request bodies to calculate responses
+by calling the handler and encoding its result.
+\begin{minted}{idris}
+(FromRequest req, ToResponse res) => HasServer (Handler req res) where
+  -- route : Handler req res -> 
+  --         (handler : IO res) ->
+  --         URL -> Maybe String -> Maybe (IO String)
+  ||| simply turn an (IO res) into an (IO String) 
+  ||| using toResponse
+  route (GET responseType) handler url requestBody =
+    Just (map (toResponse responseType) handler)
+  route (POST requestType responseType) handler url requestBody = do
+    -- see if there is a requestBody. otherwise fail
+    body <- requestBody
+    -- try to decode the request. this might fail and return Nothing
+    request <- fromRequest requestType body
+    -- apply the decoded request to the handler, and encode the result
+    Just (map (toResponse responseType) (handler request))
+\end{minted}
 
@@
 
-||| A Proof that the GET dtor is injective over its requestType
-injGET : (GET x = GET y) -> x = y
-injGET Refl = Refl
-
-||| A proof that the POST dtor is injective over its requestType
-injPOSTReq : (POST requestType1 responseType1 = POST requestType2 responseType2) -> requestType1 = requestType2
-injPOSTReq Refl = Refl
-
-||| A Proof that the POST dtor is injective over its responseType
-injPOSTResp : (POST requestType1 responseType1 = POST requestType2 responseType2) -> responseType1 = responseType2
-injPOSTResp Refl = Refl
-
-||| A Proof that a GET is not a POST
-getNotPost : (GET responseType = POST requestType responseType1) -> Void
-getNotPost Refl impossible
-
-||| A proof that GET is congruent
-cong_GET : (prf : responseType1 = responseType2) -> GET responseType1 = GET responseType2
-cong_GET = cong
-
-||| A proof POST is congruent
-cong_POST : (prf2 : responseType1 = responseType2) -> (prf1 : requestType1 = requestType2) -> POST requestType1 responseType1 = POST requestType2 responseType2
-cong_POST Refl Refl = Refl
-
-(DecEq req, DecEq res) => DecEq (Handler req res) where
-  decEq (GET responseType1) (GET responseType2) =
-    (case decEq responseType1 responseType2 of
-          (Yes prf) => Yes (cong prf)
-          (No contra) => No ((\h => contra (injGET h))))
-  decEq (POST requestType1 responseType1) (POST requestType2 responseType2) =
-    (case decEq requestType1 requestType2 of
-          (Yes prf1) => (case decEq responseType1 responseType2 of
-                             (Yes prf2) => Yes (cong_POST prf2 prf1)
-                             (No contra) => No (\h => contra (injPOSTResp h)))
-          (No contra) => No (\h => contra (injPOSTReq h)))
-  decEq (GET _) (POST _ _) = No getNotPost
-  decEq (POST _ _) (GET _) = No (negEqSym getNotPost)
-data DisjointPP : PathPart capture query -> PathPart capture query -> Type where
-  ConstD : Not (str = str') -> DisjointPP (Const str) (Const str')
-
-data DisjointPath : Path capture query req res -> Path capture query req res -> Type where
-  OutputsD : Not (handler1 = handler2) -> DisjointPath (Outputs handler1) (Outputs handler2)
-  PathBase : DisjointPP pp1 pp2 -> DisjointPath (pp1 :> p1) (pp2 :> p2)
-  PathStep : DisjointPath p1 p2 -> DisjointPath (p :> p1) (p :> p2)
-
-data DisjointAPI : List (Path capture query req res) -> Type where
-  Base : DisjointAPI []
-  Step : (x : Path capture query req res) -> All (DisjointPath x) xs -> DisjointAPI (x::xs)
 \end{minted}
@@
 
+\bibliography{sources}
+\bibliographystyle{plain}
 \end{document}

### 2016-08-22 paper/paper.tex
more text

@@
 \begin{abstract}
-In this paper, we describe a DSL for web APIs named Servis.
+\emph{TODO}
+%In this paper, we describe a DSL for web APIs named Servis.
 Servis is heavily inspired by Servant, a haskell type-level DSL for web APIs,  but tries to solve identified shortcomings by using a dependently typed language called Idris.
@@
 
-Furthermore, we take a look at if dependent types can bring more to the table in terms of expressiveness of API descriptions than the original Servant DSL\@.
-Specifically, we introduce $\pi$-types to describe interdependencies within an API description.
+%Furthermore, we take a look at if dependent types can bring more to the table in terms of expressiveness of API descriptions than the original Servant DSL\@.
+Specifically, we introduce Sigma-types to describe interdependencies within an API description.
 
-  TODO: rewrite abstract \emph{after} I finish the paper :-)
+% TODO: rewrite abstract \emph{after} I finish the paper :-)
 
@@
 \section{Introduction}
+\emph{TODO expand and reword}
+
 Servant is an extensible type-level DSL for Haskell, that tries to make web APIs \emph{first-class}~\cite{servant}.
@@
 
-TODO give a short overview of servant
+\emph{TODO give a short overview of servant}
 
@@
 
-Dependent Haskell is currently in the making, and probably planned for around GHC version 8.6~\cite{neeeded}.
+Dependent Haskell is currently in the making, and probably planned for around GHC version 8.6.
 In the future, it would probably be very feasible to port this library back to Haskell and bring all the benefits along with it.
@@
 Preferably, one would want to define a new open kind \mintinline{haskell}{API} such that \mintinline{haskell}{(:>) :: API -> API -> API}, but this is not possible in Haskell (or Idris).
-Instead, one could define a closed kind (using \textt{XDataKinds}), in which we define the entire Servant DSL.
+Instead, one could define a closed kind (using \texttt{XDataKinds}), in which we define the entire Servant DSL.
 This would mean we would lose extensibility of the DSL, but we gain kind-safety.
@@
 
+\subsection{API Interdependencies}
+\label{sec:inter}
+Sometimes, one one like to describe even more type-safety than currently possible in Servant.
+Maybe we want to let the output-type of a request depend on a query parameter.
+A use case could be the following: We have a route \texttt{"/users/list?limit=100}", and we want to make sure that a handler never returns more than 100 users.
+Thus the return type of the handler should be a \mintinline{idris}{BoundedList 100 User}.
+Or more generally, the return type should be \mintinline{idris}{BoundedList limit User}, where \mintinline{idris}{limit : Nat} comes from parsing the URL.
+The full type of the API will thus depend on some runtime value.
+This is a textbook example of dependent types, and is currently not possible to implement in Haskell.
+However, with idris this should be a breeze.
+
+\subsection{Overlapping routes}
+Another problem with Servant is that there is no protection for overlapping routes. If two routes could resolve to the same path, Servant makes a left-biased choice~\cite{servant}.
+It would be nice if we could verify at compile time that no routes in an API specification can overlap and give the user an error if he violates the overlapping constraint.
+
 \subsection{Explicit status codes}
@@
 
-It would be nice to add a list of possible status codes that an endpoint can returns, and then also be able to check whether a server implementation actually only returns one of the specified status codes if something goes wrong (or right).
-
-\subsection{Overlapping routes}
-Another problem with Servant is that there is no protection for overlapping routes. If two routes could resolve to the same path, Servant makes a left-biased choice~\cite{servant}.
-It would be nice if we could verify at compile time that no routes in an API specification can overlap.
+It would be nice to add a list of possible status codes that an endpoint can returns, and then possibly also be able to check whether a server implementation actually only returns one of the specified status codes if something goes wrong (or right).
 
@@
 Then, we will look at how we would implement HTTP webserver handlers that adhere to specification defined by the API DSL\@.
-Next, we'll look at solving the overlapping routes problem by writing proofs in Idris.
-Finally, we will extend the DSL to add \emph{pi-types}, allowing us to describe dependencies in our API\@.
+Next, we'll look at solving the overlapping routes problem by using Idris's automatic proof search.
+Finally, we will extend the DSL to add dependent types, allowing us to describe dependencies in our API\@.
 For example, one might want to describe  the size of a list of users in the API, but let the size depend on a query parameter \emph{limit} that limits the response size.
@@
 \subsection{The DSL}
-So what does this type \mintinline{idris}{API} look like?
+So what does this type \mintinline{idris}{API} look like? An API DSL should allow us to define routes, to which we can attach handler types.
 \subsubsection{Handler}\label{sec:handler}
@@
   QueryParam : (name : String) -> (type : query) -> PathPart capture query
+
+( Universe capture
+, Universe query
+) => Universe (PathPart capture query) where
+  el (Const path) = ()
+  el (Capture name type) = el type
+  el (QueryParam name type) = el type
 \end{minted}
@@
 infixr 5 :>
+
+( Universe capture
+, Universe query
+, Universe req
+, Universe resp
+) => Universe (Path capture query req resp) where
+  -- special case because we don't like () in our functions
+  el (Const path :> right) =  el right
+  el (pathPart :> right) = el pathPart -> el right
+  el (Outputs handler) = el handler
 \end{minted}
@@
 \begin{minted}{idris}
-data API : (capture : Type) -> (query : Type) -> (req : Type) -> (res : Type) -> Type where
-  OneOf : (paths : Vect (S n) (Path capture query req res)) -> API capture query req res
+data API : (capture : Type) ->
+           (query : Type) ->
+           (req : Type) ->
+           (res : Type) -> Type where
+  OneOf : (paths : Vect (S n) (Path capture query req res)) ->
+          API capture query req res
 \end{minted}
@@
 
-findUserByName : String -> IO (List User)
-getUsers : (limit : Int) -> IO (List User)
+findUserByEmail : String -> IO (List User)
+getUsers : (limit : Nat) -> IO (List User)
 getUserById : (id : Int) -> IO User
-addUser : User -> IO ()
+addUser : User -> IO User
 \end{minted}
-\label{domain}
+\label{code:domain}
 \caption{The User object and its interactions}
@@
   | QINT
+  | QNAT
 Universe QueryUniverse where
   el QSTRING = String
-  el QINT = Int
+  el QNAT = Nat
 
@@
   , Const "users" :>
-    QueryParam "limit" QINT :>
+    QueryParam "limit" QNAT :>
     Outputs (GET (LIST USER))
   ]
+
+-- set the handlers
+handlers : el userAPI
+handlers =
+  [ getUserById
+  , findUserByEmail
+  , addUser
+  , getUsers
+  ]
+
 \end{minted}
@@
 
-# /users?limit=<QINT>
+# /users?limit=<QNAT>
 ## GET
@@
 
-# /users?limit=<QINT>
-## GET
- Response type LIST USER
 \end{verbatim}
@@
 
-\subsection{A proof of non-overlapping routes}
-We will describe types for helping prove that routes in an API definition don't overlap.  We do this
-by providing Decidable Equality proofs for all parts of our DSL.
+Now we will look at how we will route a \mintinline{idris}{Path}. As we know, a path 
+is a description of how we extract information from the URL. Because parsing URLs
+is out of scope, we will assume that URLs are already parsed when they are fed into our
+routing machinery as the following datatype:
+\begin{minted}{idris}
+record URL where
+  constructor MkURL
+  pathParts : List String
+  params : List (String, String)
+\end{minted}
+
+\clearpage
+We will now give a step-by-step description of the path router.
+\begin{minted}{idris}
+( FromCapture capture
+, FromQueryParam query
+, FromRequest req
+, ToResponse resp
+) => HasServer (Path capture query req resp) where
+  -- routing a Const is simple. Just make sure the const matches
+  -- the next path part
+  route ((Const path) :> right)
+        handler
+        (MkURL pathParts params) 
+        requestBody =
+    case pathParts of
+      -- if there are no pathParts left, abort. There is no match!
+      [] => Nothing
+      (x::xs) => do
+        -- make sure the pathPart we consume matches the path part we expect
+        guard (x == path)
+        --- recurse on the remaining path parts
+        route right handler (MkURL xs params) requestBody
+
+  -- To route a capture, we pass the captured variable along to our handler
+  route ((Capture name type) :> right)
+        handler
+        (MkURL pathParts params)
+        requestBody =
+    case pathParts of
+      -- if there are no pathParts left, abort. There is no match!
+      [] => Nothing
+      (x::xs) => do
+        -- try to parse the pathPart as a capture parameter of the right type
+        -- abort if this fials
+        capture <- fromCapture type x
+        -- pass the parsed capture to handler, and recurse on
+        -- the other pathparts
+        route right (handler capture) (MkURL xs params) requestBody
+\end{minted}
+\pagebreak
+\begin{minted}{idris}
+  -- To route a query param, we look up the query param in the URL,
+  -- and pass it to the handler
+  route ((QueryParam name type) :> right)
+        handler
+        (MkURL pathParts params)
+        requestBody = do
+    -- make sure the query param that was expected exists
+    val <- lookup name params
+    -- try parse it to the appropriate type
+    param <- fromQueryParam type val
+    -- remove the parsed query param from the URL
+    let newParams = dropWhile ((== name) . fst) params
+    -- apply the param to the handler,
+    -- and recurse on the url without the query param
+    route right (handler param) (MkURL pathParts newParams) requestBody
+\end{minted}
+
+Routing multiple paths is easy. We just keep on trying consecutive paths until one
+doesn't return \mintinline{idris}{Nothing}:
+
 \begin{minted}{idris}
 
+( FromCapture capture
+, FromQueryParam query
+, FromRequest req
+, ToResponse resp) =>
+  HasServer (API capture query req resp) where
+
+  route (OneOf (path :: []))
+        (handler :: [])
+        url
+        requestBody =
+    -- if no handlers are left but this one.
+    -- We should allow it to fail
+    route path handler url requestBody
+  route (OneOf (path :: x :: xs))
+        (handler :: h :: handlers)
+        url
+        requestBody =
+    -- if other handlers are present. Try the handler
+    -- and otherwise try others
+    case route path handler url requestBody of
+      Nothing =>
+        route (OneOf (x :: xs))
+              -- NOTE: this is to help the compiler select correct type
+              (the (HVect _ ) (h :: handlers)
+              url
+              requestBody
+      a => a
 \end{minted}
-\subsection{Pi-types}
-Adds dependent pairs to our DSL such that we can describe dependencies between query params, captures and return types.  Example:
+\subsection{A proof of non-overlapping routes}
+\emph{TODO}
+\subsection{Adding API interdependencies}
+As discussed in Section~\ref{sec:inter}, it would be nice to describe more complexly typed APIs, where we allow interdependencies within an API path.
 
-\begin{minted}{haskell}
-exampleOfDependentPath : Path EmptyU QueryU EmptyU RespU
-exampleOfDependentPath =
-  QueryParam "limit" NAT :*> \limit => Outputs (GET (BLIST limit USER))
+We can easily add this functionality to our DSL by introducing a \emph{dependent} version of \mintinline{idris}{:>}, named
+\mintinline{idris}{:*>} in the \mintinline{idris}{Path} datatype.
+What we want is that that the value of the lefthand-side of \mintinline{idris}{:*>} influences the type of the right-hand side.
+\begin{minted}{idris}
+data Path : capture -> query -> req -> res -> Type where
+  Outputs : (handler : Handler req res) -> Path capture query req res
+  (:>) :  (pathPart : PathPart capture query) ->
+          (path : Path capture query req res) -> Path capture query req res
+  -- The RHS of :*> now takes a function instead of a path. that calculates a path
+  (:*>) : (pathPart : PathPart capture query) ->
+          (path : (el pathPart -> Path capture query req res)) ->
+          Path capture query req res
+\end{minted}{idris}
+This is enough to express basic dependent APIs.
 
-getUsersLimit : (limit : Nat) -> IO (BoundedList limit User)
-getUsersLimit limit = pure . take limit $ users
+A good usecase for these kind of dependent types is for endpoints that return a \mintinline{idris}{BoundedList}.
+A bounded list in idris is a list that can only contain upto a certain amount of elements. It is defined as follows:
+\begin{minted}{idris}
+data BoundedList : Nat -> Type -> Type where
+  Nil : BoundedList n a
+  (::) : a -> BoundedList n a -> BoundedList (S n) a
+\end{minted}
 
-exampleOfDependentPathHandler : el Example.exampleOfDependentPath
-exampleOfDependentPathHandler = getUsersLimit
+Lets extend our Universe of Listing~\ref{code:api} to include bounded lists:
+\begin{minted}{idris}
+data ReqResUniverse
+  = USER
+  | LIST ReqResUniverse
+  | BLIST Nat ReqResUniverse
+Universe ReqResUniverse where
+  el USER = User
+  el (LIST a) = List (el a)
+  el (BLIST n a) = BoundedList n (el a)
+\end{minted}
+And lets modify the domain function that queries the database for users
+in Listing~\ref{code:domain} to return bounded lists:
 
+\begin{minted}{idris}
+getUsers : (limit : Nat) -> IO (BoundedList limit User)
 \end{minted}
-\subsection{New problems that pi-type create}
-Here we will describe how pi-types adds problems. For example we cannot easily generate documentation anymore, prove that there are no overlapping routes, or prove that statuscodes are adhered to.
+We can now describe an API in which we can describe the dependency between the query parameter
+and the size of the returned list of users.
+\begin{minted}{idris}
+getUsersPath : Path CaptureUniverse QueryUniverse ReqResUniverse ReqResUniverse
+getUsersPath = Const "users" :> 
+              QueryParam "limit" QNAT :*>
+              \limit => Outputs (GET (BLIST limit USER))
 
-\section{Summary}
+handler : el getUsersPath
+handler = getUsers
+\end{minted}
+
+Next, we will need to add extra cases to the \mintinline{idris}{HasServer}
+implementation of \mintinline{idris}{Path} to handle the dependency. We
+will only show the implementation for \mintinline{idris}{QueryParam},
+but the changes to make \mintinline{idris}{Capture} work are identical.
+
+The idea is, that instead of only passing the parsed query parameter value
+to the handler, we also pass it along to the right hand side of \mintinline{idris}{:*>}:
+\begin{minted}{idris}
+  route ((QueryParam name type) :*> right)
+        handler
+        (MkURL pathParts params)
+        requestBody = do
+    val <- lookup name params
+    param <- fromQueryParam type val
+    let newParams = dropWhile ((== name) . fst) params
+    -- BEFORE :
+--  route right 
+--        (handler param)
+--        (MkURL pathParts newParams)
+--         requestBody
+    -- WITH ADDED DEPENDENCY:
+    route (right param) -- <- note the extra argument passed to RHS
+          (handler param)
+          (MKURL pathParts newParams)
+          requestBody
+\end{minted}
+And that's it. we now have added dependent types to our API!
+
+\subsection{New problems that dependent types create}
+The dependent types are nice, but our datatype is now a Higher-order abstract syntax instead of a first-order abstract syntax.
+Which means the datatype contains a function.  This makes it difficult to generate documentation as we need to provide a value
+to apply to the function to traverse the datatype. It isn't impossible though. We could introduce the following interface:
+\begin{minted}{idris}
+interface Universe u => HasSample u where
+  sample : (v : u) -> el v
+
+data ExampleU = INT | STRING
+
+Universe ExampleU where
+  el INT = Int
+  el STRING = String
+
+HasSample ExampleU where
+  sample INT = 0
+  sample STRING = "A string"
+\end{minted}
+Now in our implementation of \mintinline{idris}{HasDocs} we can just call \mintinline{idris}{sample}
+every time we encounter a function:
+\begin{minted}{idris}
+HasDocs (Path capture query req res) where
+  docs ((Capture str type) :*> f) =
+    "super nice docs"  ++ docs (f (sample type))
+\end{minted}
+
+The real problem though is that this representation breaks the overlapping path detector.
+The overlapping path detector uses the feature of idris to search proofs automatically. But that
+proof search isn't perfect, and won't be able to prove that paths overlap or not if we have
+arbitrary functions in our datatypes.
+
+A solution this problem would be that library authors would have to provide their own proof
+of whether or not their api definitions overlap, but this can be very cumbersome and thus
+is undesired.
+
 \section{Related work}
-Talk about tel/serv. An inspiration for this project.
+This project was heavily inspired by \url{https://github.com/tel/serv}. Which is
+a closed-kinded Servant alternative. When I stumbled upon it, I got very curious
+what would happen if we would port it to Idris.  The serv source code is kind of
+dense because Haskell isn't a very nice language to do type-level programming in
+and it got me curious whether or not idris would improve readability. Which it did!
+
+Also, once I had a working example of Serv in idris, I wanted to see if idris
+could add any new tricks to it. This made me implement dependent types and
+route conflict resolution.
+
 \section{Conclusions and future work}
+Idris allows us to elegantly describe APIs with plain ol' datatypes. We
+can then use the Universe pattern to map these datatypes to handler
+function types. These handler function types can then be made routable
+by writing an interpreter of the API datatype that decodes the incoming
+requests and passes the correct parameters to the handlers.
 
+Idris makes it easier to write utility functions, like documentation generation
+or code generation, because our API definition does not live in the type-level
+anymore, because type-level and term-level are united in idris.
+
+We were able to solve the common mistake of overlapping routes in Servant as well,
+by using Idris's proof search capabilities.
+
+Finally, we were able to implement an dependent API, that allows interdependencies
+within paths to be able to describe even more powerful APIs.
+
+In the future, it would be interesting to look at if it's possible to enforce in 
+the type-system that handlers only return HTTP Status codes that are defined
+in the API description. I explored some ideas on how to implement this, but
+did not have the time to do an implementation of this.
 \bibliography{sources}

### 2016-08-25 paper/paper.tex
Write more and more

@@
 \usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
+\makeatletter
+\renewcommand*{\thelisting}{\thesection.\arabic{listing}}
+\@addtoreset{listing}{section}
+\makeatother
 \author{Arian van Putten \\ a.vanputten@uu.nl}
@@
 \begin{abstract}
-\emph{TODO}
-%In this paper, we describe a DSL for web APIs named Servis.
-Servis is heavily inspired by Servant, a haskell type-level DSL for web APIs,  but tries to solve identified shortcomings by using a dependently typed language called Idris.
-Servis uses the universe pattern to deduce types from an API description such that we can verify at compile time that the implementation of the API matches the description.
-
-%Furthermore, we take a look at if dependent types can bring more to the table in terms of expressiveness of API descriptions than the original Servant DSL\@.
-Specifically, we introduce Sigma-types to describe interdependencies within an API description.
-
-% TODO: rewrite abstract \emph{after} I finish the paper :-)
+In this paper we describe a DSL for web APIs named Servis.
+Servis is  inspired by Servant, a haskell type-level DSL for web APIs.
+We identify several issues with the current implementation of Servantm and show how Servis solves these by using features of the dependently typed language called Idris.
+
+Specifically, we show how dependent types can help use describe dependencies within an API, how we can check at  compile time that routes in an API dont overlap, and how dependent pairs can let us verify that status codes described in an API are adhered to.
 
@@
 \section{Introduction}
-\emph{TODO expand and reword}
-
 Servant is an extensible type-level DSL for Haskell, that tries to make web APIs \emph{first-class}~\cite{servant}.
@@
 
-It does this by describing the DSL at the type-level.
-And then uses \emph{type families} (type-level functions) to be able to perform computations on the API specifications.
-Using these type families, we are able to write different \emph{interpretations} of an API\@.
+It does this by describing the DSL at the type-level, and then uses \emph{type families} (type-level functions) to be able to perform computations on the API specifications.
+By using these type families, we are able to write different \emph{interpretations} of an API\@.
 Example of interpretations are: generating documentation from a specification, generating types of a server implementation, or even generating a client that can interfact with the API\@.
 
-In this paper, I will describe a port of Servant in Idris, and explore what kind of benefits this brings us.
+In this paper, We will describe Servis, a DSL  similar to  Servant written in Idris. We will point out various issues with the current Servant DSL,
+and then show how we can solve this in Idris.
 
 \subsection{An introduction to Servant}
-I will now give a very short overview of the internals of Servant. We will start by describing the type-level DSL of servant.
+\label{sec:servant_intro}
+We will now give a very short overview of the internals of Servant. We will start by describing the type-level DSL of servant.
+
+Servant lives in the type-level, and defines its language as part of the kind of types named \mintinline{idris}{*}, which is an open kind.
+This has been done on purpose so that the language becomes extensible.
+We can always add new constructs to the DSL by creating new data definitions.
 
@@
 data JSON
-\end{minted}
-\caption{The Servant DSL}
+data PlainText
+
+type Echo = "echo"
+:> ReqBody ’[PlainText] String
+:> Get
+’[PlainText] String
+\end{minted}
+\caption{The Servant DSL and an example of an API definition}
 \label{lab:servant}
@@
 
-\emph{TODO give a short overview of servant}
-
-Servant has a few limitations which we will describe in Section~\ref{sec:limitations}.
+It then uses type classes and type families to describe operations on this type-level DSL. For example,
+a documentation generation type class exists:
+\begin{minted}{haskell}
+docs :: HasDocs api => Proxy api -> String
+\end{minted}
+
+A more exotic example is the server type class, that transforms an API into a web server:
+\begin{minted}{haskell}
+class HasServer api where
+  type Server api :: *
+  route :: Proxy api -> Server api -> RoutingApplication
+
+serve :: HasServer api => Proxy api -> Server api -> Application
+
+echo :: Server Echo 
+echo txt = return txt
+
+main :: IO ()
+main = run 8080 (serve (Proxy Echo) echo)
+\end{minted}
 
@@
 Idris is a dependently-typed language.
-This means that types are fist-class citizens.
+This means that types are first-class citizens.
 We can pass types to functions, let functions create new types, and let types depend on other value-level terms.
 
-Though haskell has type-level functions (called type-families), it does not allow these functions to dpend on term-level values.
+Though haskell has type-level functions (called type-families), it does not allow these functions to depend on term-level values.
 There is a distinct boundary between type-level and term-level, whilst the two are merged in Idris.
-This makes doing computations with types a lot more natural (no special constructs are needed, just write functions), but also lets us do some more powerful stuff like letting types actually depend on runtime user input.
+This makes doing computations with types a lot more natural.
+No special constructs are needed like special syntax for type-families; you can just write functions like you would normally do with term-level values.
+It also allows us to do more powerful things like letting types actually depend on runtime user input.
 
@@
 \begin{itemize}
-  \item I point out several problems an limitations of the original Servant DSL
-  \item I show a port of servant to Idris, a dependently-typed language
-  \item I show how new techniques from Idris can help mitigate some of the problems that Servant has.
+  \item We point out several problems and limitations of the original Servant DSL
+  \item We show a port of servant to Idris, a dependently-typed language
+  \item We show how new techniques from Idris can help mitigate some of the problems that Servant has.
 \end{itemize}
@@
 As explained in the introduction, the Servant DSL lives in the type-level.
-In haskell, types and terms are separated.
-One can lift terms into the type-level using GHC extensions like \texttt{XDataKinds}, but that only gets us so far.
+In Haskell, types and terms are separated.
+One can lift terms into the type-level using GHC extensions like \texttt{XDataKinds}.
 On the term level, Haskell has many great libraries and data structures that can be used.
@@
 It would be useful to be able to use these functions to manipulate APIs, making them truly first-class.
-Now they are only first-class in the type level.
-
-For example, a useful function would be \mintinline{haskell}{path "/users/friends/etc" = "users" :> "friends" :> "etc"} to reduce noise in the DSL\@.
-But this is currently impossible to write because Strings in the type level are of kind \mintinline{haskell}{Symbol} and not \mintinline{haskell}{String} so all our favorite string manipulation functions like \mintinline{haskell}{splitOn "/"} will not work.
+But there is no way to lift existing term-level functionality to the type-level, which makes an API description only first-class on the type level.
+
+For example, a useful function would be \\ \mintinline{haskell}{path "/users/friends/etc" = "users" :> "friends" :> "etc"} to reduce noise in the DSL\@.
+But this is currently impossible to write because Strings in the type level are of kind \mintinline{haskell}{Symbol} and not \mintinline{haskell}{String} so useful string manipulation functions like \mintinline{haskell}{splitOn "/"} will not work.
 This would mean that we would have to re-implement all kinds of utility functions on the type-level.
@@
 \subsection{A kind of APIs}
-As mentioned in Section~\ref{sec:servant_intro}, a delibirate choice has been made to let the Servant DSL live in the open kind \mintinline{haskell}{*}.
+As mentioned in Section~\ref{sec:servant_intro}, a deliberate choice has been made to let the Servant DSL live in the open kind \mintinline{haskell}{*}.
 Though this makes the DSL extensible, it also makes the DSL fragile and prone to error.
@@
 
-The problem is hard to spot, but the user typed wrote \mintinline{haskell}{->} instead of \mintinline{haskell}{:>}.
+The problem is hard to spot, but the user wrote \mintinline{haskell}{->} instead of \mintinline{haskell}{:>}.
 This code compiles fine though.
 Because the Servant DSL lives in kind \mintinline{haskell}{*}, \mintinline{haskell}{:>} is of kind \mintinline{haskell}{* -> * -> *}.
-Which is the same kind as the kind of \mintinline{haskell}{->}.
+The type constructor of functions, \mintinline{haskell}{->} is also of kind \mintinline{haskell}{* -> * -> *}, so there is no way for the compiler to tell that we used the wrong operator here.
 
@@
 
-Also, as described in Section~\ref{sec:servant_intro}, the Servant DSL is split up in several semantic portions.
+Also, as seen in Listing~\ref{lab:servant}, the Servant DSL is split up in several semantic portions.
 For example, \mintinline{haskell}{:>} should take an \mintinline{haskell}{item} on the left-hand side, which describes either a constraint like \mintinline{haskell}{ReqBody} or a path-piece, whilst the right-hand side should be an \mintinline{haskell}{api}.
-These semantic portions are implict and not checked though.
-So we can create totally non-sensical API types like \mintinline{haskell}{type Nonsense = (Int :>Int) :<|> ("hey ":> Int)} that have no interpretation at all.
+These semantic portions, however,  are implict and unchecked, so we can create totally non-sensical API types like \mintinline{haskell}{type Nonsense = (Int :>Int) :<|> ("hey ":> Int)} that have no interpretation at all.
 
@@
 
-Another implict rule is, that a chain of \mintinline{haskell}{:>} should always end in a \mintinline{haskell}{Get} or a \mintinline{haskell}{Post}.
+Another implict rule is that a chain of \mintinline{haskell}{:>} should always end in a \mintinline{haskell}{Get} or a \mintinline{haskell}{Post}.
 So this should be invalid: \mintinline{haskell}{type API = "a" :> ReqBody [JSON] User}.
@@
 \label{sec:inter}
-Sometimes, one one like to describe even more type-safety than currently possible in Servant.
-Maybe we want to let the output-type of a request depend on a query parameter.
-A use case could be the following: We have a route \texttt{"/users/list?limit=100}", and we want to make sure that a handler never returns more than 100 users.
+Sometimes, one would like to describe even more type-safety than is currently possible in Servant. Maybe we want to let the output-type of a request depend on a query parameter.
+A usecase could be the following: We have a route \texttt{"/users/list?limit=100}", and we want to make sure that a handler never returns more than 100 users.
 Thus the return type of the handler should be a \mintinline{idris}{BoundedList 100 User}.
@@
 This is a textbook example of dependent types, and is currently not possible to implement in Haskell.
-However, with idris this should be a breeze.
+However, with Idris this does become possible.
 
 \subsection{Overlapping routes}
-Another problem with Servant is that there is no protection for overlapping routes. If two routes could resolve to the same path, Servant makes a left-biased choice~\cite{servant}.
-It would be nice if we could verify at compile time that no routes in an API specification can overlap and give the user an error if he violates the overlapping constraint.
-
-\subsection{Explicit status codes}
-An API that describes how stuff goes right is good, and API that describes how stuff goes wrong is great.
-But currently, Servant has no way to indicate in the API description what kind of status codes an endpoint can return.
-Thus this also does not show up in the generated documentation.
-
-It would be nice to add a list of possible status codes that an endpoint can returns, and then possibly also be able to check whether a server implementation actually only returns one of the specified status codes if something goes wrong (or right).
+Another problem with Servant is that there is no protection for overlapping routes.
+If two routes resolve to the same path, Servant makes a left-biased choice~\cite{servant}.
+It would be useful that instead of making a design choice where some routes are shadowed without warning, the compiler could inform the user about a potential mistake.
 
@@
          :<|> "users" :> "favorites" :> GET [JSON] User
-\end{minted}{haskell}
+\end{minted}
 \caption{An example of overlapping routes. The capture will be chosen if args equals "favorites"}
 \end{listing}
+
+\subsection{Explicit status codes}
+It is great that Servant lets us give a presice description about the structure and types of our web APIs. However, it does now allow us to describe any form of errors within the API types.
+It is important for a developer to know how an API call could fail, such that he can handle it sufficently.
+Also, when generating documentation from an API type, it one would like to have this information at hand to inform the users of your API on what kind of errors they can expect.
+
+This would not be hard to add to Servant.
+A type-level list of error types paired with the return type of an API endpoint would suffice.
+However, it would also be interesting to explore if we can enforce that the code that implements a server for an API, actually only returns the error codes that are documented.
 
@@
 It tries to address each issue listed in Section~\ref{sec:issues}.
-Firstly, we will define the DSL in which we can write API specifications.
-Then, we will look at how to write an interpreter of this DSL that generates documentation.
+Firstly, we will define the DSL in which we can write API specifications. Then, we will look at how to write an interpreter of this DSL that generates documentation.
 Then, we will look at how we would implement HTTP webserver handlers that adhere to specification defined by the API DSL\@.
@@
 
- In idris, we could define this as the following interface:
+ In Idris, we can define this as the following interface:
  \begin{listing}
@@
   el : u -> Type
- \end{minted}{idris}
+ \end{minted}
    \caption{Universe interface in Idris}
@@
 \subsubsection{Handler}\label{sec:handler}
-Before we will look at routes, lets think of what an endpoint should look like.
+Before we will look at routes, we must decide what an endpoint should look like.
 We should be able to handle the HTTP verbs. We will only implement \mintinline{idris}{GET} and \mintinline{idris}{POST}, because the others are very similar.
 For a \mintinline{idris}{GET}, we should be able to specify a return type of the handler,  and for a \mintinline{idris}{POST}, we should be able to
-specify a request body type, and a return type.
+specify a request body type and a return type.
 
@@
 
-However, once we start implementating an actual server in Section~\ref{sec:server}, we will discover that this definition is not ideal. What if we want to write some
-function that does something different based on the \mintinline{idris}{responseType} of a \mintinline{idris}{GET}?
-For example, based on value of \mintinline{idris}{responseType}, we need to decide how to encode the response body to JSON.
-Lets imagine a hypothetical function \mintinline{idris}{selectEncoder} that tries to implement this behaviour:
+However, once we start implementing an actual server in Section~\ref{sec:server}, we will discover that this definition is not ideal.
+What if we would want to write a function that needs to decide its behaviour based on the \mintinline{idris}{responseType} of a \mintinline{idris}{GET}?
+For example, we have a piece of functionality in our code base, where we need to decide how to encode a response body to JSON, based on the value of \mintinline{idris}{responseType}.
+Lets try to implement such a function \mintinline{idris}{selectEncoder} and see what road blocks we hit:
 
@@
 userEncoder = selectEncoder User
-\end{minted}{idris}
-
-Sadly enough, this will give us a compiler error. The problem is that in Idirs, we cannot pattern match on values of \mintinline{idris}{Type}. 
-However, this exact problem can be solved with universes! We can define a data type \mintinline{idris}{ResponseUniverse : Type} which names
-the \mintinline{idris}{User} type. We can then pattern-match on \mintinline{idris}{ResponseUniverse} instead of \mintinline{idris}{Type}!
-And whenever we need the \mintinline{idris}{User} type, we can just call \mintinline{idris}{el USER : Type}.
+\end{minted}
+
+Sadly enough, this code will give us a compiler error. The problem is that in Idirs, we cannot pattern match on values of \mintinline{idris}{Type}. 
+However, this exact problem can be solved with universes We can define a data type \mintinline{idris}{ResponseUniverse : Type} which names
+the \mintinline{idris}{User} type as a data constructor \mintinline{idris}{USER : ResponseUniverse}.
+We can then pattern-match on \mintinline{idris}{ResponseUniverse} instead of \mintinline{idris}{Type}, and whenever we need the \mintinline{idris}{User} type, we can just call \mintinline{idris}{el USER : Type}.
 
@@
 
-This leads us to the following definition of \mintinline{idris}{Handler}:
-
+We now change the definition of \mintinline{idris}{Handler} to be parameterised over a request universe (\mintinline{idris}{req : Type}) and a response universe (\mintinline{idris}{res : Type}):
 \begin{listing}
@@
 is where \mintinline{idris}{Path}s and \mintinline{idris}{PathPart}s come into play. 
-When we talk about path parts, we mean sections of an path template that convey parsable information.
+When we talk about path parts, we mean sections of a path template that convey parsable information.
 For example, the following path template has three path parts: \texttt{/users/\string{user\_id\string}?search=\string{query\string}}. A constant piece
@@
 
-||| PathPart is a part of a path.
-||| @ capture the universe of captures
-||| @ query   the universe of query params
 data PathPart : (capture : Type) -> (query : Type) -> Type where
@@
 
-The Path we just described can now be written as
+The path \texttt{/users/\string{user\_id\string}?search=\string{query\string}}, can be described as follows with these new data types:
 \begin{minted}{idris}
@@
 The universe implementation is straightforward for paths, except for
-that we do not wannt to have spurious arguments in case when we encounter
+that we do not want to have spurious arguments in case when we encounter
 a \mintinline{idris}{Const}: 
@@
 \subsubsection{The API Datatype}
-Finally, we define an \mintinline{idris}{API} as an non-empty list of \mintinline{idris}{Path}s.
+Finally, we define an \mintinline{idris}{API} as an non-empty list of \mintinline{idris}{Path}s that can be chosen from.
 
@@
 is of type \mintinline{idris}{el x}.
-\begin{minted}
+\begin{minted}{idris}
 ( Universe capture
@@
   el (OneOf xs) = HVect (map el xs)
-\end{minted}.
+\end{minted}
 
@@
 Using this DSL, we will now describe an API for interacting with users in a database.
-We will start of with defining our domain objects, and intercations we could do with them.
-
-\begin{listing}
+We will start off with defining our domain objects, and interactions we can perform on them.
+These definitions can be found in Listing~\ref{code:domain}.
+
+\begin{listing}[H]
 \begin{minted}{idris}
@@
 getUsers : (limit : Nat) -> IO (List User)
-getUserById : (id : Int) -> IO User
 addUser : User -> IO User
@@
 Next, we will describe the HTTP API for interacting with these users.  This includes
-the need of coming up with universes for captures, query params, request body, and response.
-
-\begin{listing}
+the need of coming up with universes for query params, request body, and response. We will use this example in the upcoming sections
+to show the behaviour of different interpreters.
+
+% TODO MAKE A GOOD PLACEMENT FOR THIS AND MAYBE SHORTER.
+% TODO DESCRIBE EACH STEP. BREAK IT UP!!
+\begin{listing}[H]
 \begin{minted}{idris}
@@
 
-data CaptureUniverse
-  = CINT
-Universe CaptureUniverse where
-  el CINT = Int
-
 -- Note: We use the same universe for requests and responses.
@@
 userAPI = OneOf
-  [ Const "users" :> Capture "user_id" CINT :> Outputs (GET USER)
-  , Const "users" :>
+  [ Const "users" :>
     Const "find" :>
@@
 handlers =
-  [ getUserById
-  , findUserByEmail
+  [ findUserByEmail
   , addUser
@@
 \subsection{A simple interpreter: Documentation generation}
-Writing an interpreter for generating documentation is very easy.
-Our API description isn't some type-level definition, it's just a plain old datatype that can be passed to functions and can be manipulated.
-
-We start by introducing an interface for docs:
+Writing an interpreter for generating documentation is straightforward.
+Our API description isn't some type-level definition, it's just a datatype that can be passed to functions and can be manipulated.
+
+We start by introducing an interface for docs, which just tells us how to turn something into a string:
 \begin{minted}{idris}
@@
 \end{minted}
-Now, we just define an implementation for \mintinline{idris}{API} for this interface and we're done. Which is just calling \mintinline{idris}{docs} recursively on all its children.
-
+Now, we just need to define an implementation for \mintinline{idris}{API} for this interface and we are done.
+To implement this, we will of course also needs to implement the interface for \mintinline{idris}{Handler}, \mintinline{idris}{PathPart} and \mintinline{idris}{Path}.
+
+To get an idea on how this would work, we will show the implementation of \mintinline{idris}{Handler} and \mintinline{idris}{API}:
 \begin{minted}{idris}
@@
     "Response type: " ++ docs responseType ++ "\n"
-
-renderParams : HasDocs query => List (String, query) -> String
-renderParams = renderParams' . reverse
-  where
-    renderParams' : List (String, query) -> String
-    renderParams' [] = ""
-    renderParams' ((a, b) :: []) =
-      "?" ++ a ++ "=<"++ docs b ++ ">"
-    renderParams' ((a, b) :: xs) =
-      renderParams' xs ++ "&" ++ a ++ "=<"++ docs b ++ ">"
-
-renderPath : 
-  ( HasDocs capture
-  , HasDocs query
-  , HasDocs req
-  , HasDocs resp) => Path capture query req resp -> String -> String
-renderPath (Outputs x) params = 
-    params ++ "\n" ++ docs x
-renderPath ((Const x) :> y) params =
-    "/" ++ x ++ renderPath y params
-renderPath ((Capture name type) :> y) params =
-  "/<" ++ name ++ ":" ++ docs type ++ ">" ++ renderPath y params
-renderPath _ _ = ""
-
-( HasDocs capture
-, HasDocs query
-, HasDocs req
-, HasDocs resp
-) => HasDocs (Path capture query req resp) where
-  docs path =
-    "# " ++ renderPath path (renderParams . getParams $ path) ++ "\n"
-    where
-
-      getParams : Path capture query req resp -> List (String, query)
-      getParams (Outputs x) = []
-      getParams ((Const x) :> y) = getParams y
-      getParams ((Capture name type) :> y) = getParams y
-      getParams ((QueryParam name type) :> y) = (name, type) :: getParams y
-
+\end{minted}
+
+% TODO add text
+\begin{minted}{idris}
 ( HasDocs capture
@@
 \begin{verbatim}
-# /users/<user_id:CINT>
-## GET
- Response type: USER
 # /users/find?email=<QSTRING>
@@
 \subsection{An interpreter for an HTTP Server}\label{sec:server}
-Now that we've shown that it is easy to manipulate APIs because they're just datatypes, it's time for a more complicated example.
-Though it is cool that we can generate documentation from an API description, the useful part that we want to implement is an HTTP Server.
+
+Now that we have shown a small example on how to do computations on API descriptions,
+it is time for a less contrived example. We would like to implement handlers for our API and then serve these on the web.
 
 \subsubsection{A type for web applications}
-Lets first ask the question how a web server should look like in Idris. In the Servant paper, they describe the type of http handlers in the \texttt{wai} library, which looks like \cite{servant}:
+Before we write such an interpreter, lets first ask the question what a web server should look like in Idris.
+In the Servant paper, the authors describe the type of http handlers in the \texttt{wai} library, which looks like \cite{servant}:
 \begin{minted}{haskell}
@@
 \end{minted}.
-
-We will more-or-less use this representation to model our HTTP handlers.
-they are  simply a function from request to response.
-The actual implementation of a web server based on this model is out of scope for now, but it would be interesting to port something like Haskells' \texttt{warp} (which impelements this concept) to Idris.
+This is a natural way to think of an HTTP application. A request comes in, and we apply a function of type \mintinline{idris}{Application} to get a response, which can then be sent back.
+
+We will more-or-less use this representation to model our HTTP handlers.An actual implementation of a web server based on this model is out of scope for now, but it would be interesting to port something like Haskells' \texttt{warp} (which impelements this concept) to Idris.
 
 Instead of having two abstract types \mintinline{idris}{Request} and \mintinline{idris}{Response}, we will assume for simplicity that a \mintinline{idris}{Response} is just a \mintinline{idris}{String}, and that a \mintinline{idris}{Request} is an \mintinline{idris}{URL}, and a requestBody of type \mintinline{idris}{Maybe String}.
+This leads us to the following definition for \mintinline{idris}{Application}:
 
@@
 \end{minted}
-
-Just like in the Servant paper, we will need to modify this type a little bit more to suit our needs.
-We need to be able to express that an \mintinline{idris}{Application} does not match in case of a routing mismatch, such that we can try another route encompassed in \mintinline{idris}{OneOf}.
-
-Instead of returning  a value of \mintinline{idris}{IO String}, we can return a value of \mintinline{idris}{Maybe (IO String)}, where \mintinline{idris}{Just} signals a route match, whilst \mintinline{idris}{Nothing} signals a route mistmatch such that we can try another route.
+% TODO reword a bit
+
+The router we are going to build uses the same mechanism as described in the Servant paper. In order to be able to write a router, we will have
+to modify the definition of \mintinline{idris}{Application} slightly~\cite{servant}. We need to be able to express that the router can not
+match the API path with the incoming request. In case of a mismatch, the router should then try if the next available path does match the request.
+Instead of letting the application return a value of \mintinline{idris}{IO String}, let it return a value of \mintinline{idris}{Maybe (IO String)}, where \mintinline{idris}{Just a} signals a route match, whilst \mintinline{idris}{Nothing} signals a route mistmatch such that we can try another route.
 
@@
 \subsubsection{Decoding and encoding data}
-In section~\ref{sec:handler}, we showed that if we want to implement some decoder, we would have to resort
-to the universe pattern. So that is what we will do exactly here.
-
-As an example, we will show how we would decode a queryparam. All other decoders follow the same pattern.
+In Section~\ref{sec:handler}, we showed that if we want to implement some decoder, we would have to resort to the universe pattern.
+We will use this technique here for implementing decoding of requests, and encoding of responses.
+
+As an example, we will show how we would decode a query parameter. All other decoders follow the same pattern.
 What we want to do is, given a \mintinline{idris}{queryParam : String}, and a value \mintinline{idris}{v : u},
@@
 \end{minted}
-
-similarly, we have interfaces for captures and request bodies:
+Similarly, we have interfaces for captures and request bodies:
 \begin{minted}{idris}
@@
   fromCapture : (v : u) -> String -> Maybe (el v)
-
 interface Universe u => FromRequest u where
@@
 \end{minted}
-
-An implementation of \mintinline{idris}{FromQueryParam} for the universe \mintinline{QueryUniverse} in Listing~\ref{code:api}, would look like this:
-
-\begin{minted}{idris}
-FromQueryParam QueryU where
+An implementation of \mintinline{idris}{FromQueryParam} for the universe \mintinline{idris}{QueryUniverse} in Listing~\ref{code:api}, would look like this:
+\begin{minted}{idris}
+FromQueryParam QueryUniverse where
   fromQueryParam QINT str = maybeParseInt str
 \end{minted}
-
-The fact that we can pattern match on the universe lets us inspect what type we need to return,
-and select the right decoder accordingly.
+The fact that we can pattern match on the universe lets us inspect what type we need to return, and select the right decoder accordingly.
 
@@
   toResponse : (v : u) -> el v -> String
-
-\end{minted}
-
+\end{minted}
+% TODO reword
+% TODO CHECKED UNTIL HERE. PAGE 11
 For example, we could use this to encode values in the \mintinline{idris}{ReqResUniverse} universe from
@@
 \subsubsection{Routing requests}
-We will define the following interface to implement routing requests:
+To route a request, we should be able to pattern match on our API, (which is a universe), and apply data from the request
+to a handler function whose type is dependent on the API universe. This leads us to the following definition for a router:
 \begin{minted}{haskell}
@@
 \end{minted}
-Or if we expaned the definition \mintinline{idris}{RoutingApplication}:
+Or if we expand the definition of \mintinline{idris}{RoutingApplication}:
 \begin{minted}{haskell}
@@
           (requestBody : Maybe String)
-          -> Maybe (IO String)
-\end{minted}
-
-The idea of the router is, that given some part of the API \mintinline{idris}{v : u}, and a handler of type \mintinline{idris}{el v}, build an application that can either route succesfully or fail.
-Preferably, the application uses the handler parameter in its implementation of course.
-
-Lets start by defining an \mintinline{idris}{HasServer} instance for \mintinline{idris}{Handler}. Its task is to simply encode responses to Strings, and to handle request bodies to calculate responses
-by calling the handler and encoding its result.
+          -> RouteResult (IO String)
+\end{minted}
+
+The first part of the API for which we will implement the routing mechanics is the \mintinline{idris}{Handler}.
+Defining this router consists of two cases: A \mintintline{idris}{GET} or a \mintinline{idris}{POST}.
+During a \mintinline{idris}{GET}, we just have to execute the handler which will then give us a response. 
+For a \mintinline{idris}{POST}, we need to read the request body from the request, decode it to the right type, and
+apply it to the handler to get a response:
 \begin{minted}{idris}
@@
   -- route : Handler req res -> 
-  --         (handler : IO res) ->
+  --         (handler : IO (el res)) ->
   --         URL -> Maybe String -> Maybe (IO String)
-  ||| simply turn an (IO res) into an (IO String) 
+  ||| simply turn an (IO (el res)) into an (IO String) 
   ||| using toResponse
@@
 
-Now we will look at how we will route a \mintinline{idris}{Path}. As we know, a path 
+Next up is routing a \mintinline{idris}{Path}. A path 
 is a description of how we extract information from the URL. Because parsing URLs
@@
 
-\clearpage
 We will now give a step-by-step description of the path router.
-\begin{minted}{idris}
-( FromCapture capture
-, FromQueryParam query
-, FromRequest req
-, ToResponse resp
-) => HasServer (Path capture query req resp) where
-  -- routing a Const is simple. Just make sure the const matches
-  -- the next path part
+We can check if this URL matches the \mintinline{idris}{Path} description, and extract
+the appropriate query parameters and captures.
+
+In case of a \mintinline{idris}{Const}, we will have to check if our URL starts with the
+same part as the name of the \mintinline{idris}{Const}:
+
+\begin{minted}{idris}
   route ((Const path) :> right)
@@
         route right handler (MkURL xs params) requestBody
-
-  -- To route a capture, we pass the captured variable along to our handler
+\end{minted}
+In case of a capture, we almost do the same thing, but instead of trying to match
+the path parts name, we try to parse it as the appropriate type using \mintinline{idris}{fromCapture}:
+\begin{minted}{idris}
   route ((Capture name type) :> right)
@@
     case pathParts of
-      -- if there are no pathParts left, abort. There is no match!
       [] => Nothing
@@
 \end{minted}
-\pagebreak
-\begin{minted}{idris}
-  -- To route a query param, we look up the query param in the URL,
-  -- and pass it to the handler
+In case of a query parameter, we do a lookup in the URL if the expected
+query parameter was present, and then try to parse it as the appropriate type:
+\begin{minted}{idris}
   route ((QueryParam name type) :> right)
@@
 Routing multiple paths is easy. We just keep on trying consecutive paths until one
-doesn't return \mintinline{idris}{Nothing}:
-
-\begin{minted}{idris}
-
+does nott return \mintinline{idris}{Nothing}:
+\begin{minted}{idris}
 ( FromCapture capture
@@
 \end{minted}
+
+\subsection{Explicit status codes}
+To better document the behaviour of our API to users of the API, we can add a list of status codes that an endpoint can return by modifing \mintinline{idris}{Handler}:
+
+\begin{minted}{idris}
+data Handler : (req : Type)  -> (res : Type) -> Type where
+  GET : (responseType : res) -> (statuses : List Int) -> Handler req res
+  POST : (requestType : req) -> (responseType : res) ->
+         (statuses : List Int)  -> Handler req res
+\end{minted}
+
+However, it would be convenient if we could automatically deduce a proof that our handlers never return
+other status codes than are in that list. However, this gets impossible quickly because not all programs are decidable.
+Instead, we can let our \mintinline{idris}{Universe} implementation generate a type that forces the implementor of the API to provide a proof for us.
+We can accomplish this by letting the universe implementation return a dependent pair.
+A dependent pair in idris is a tuple where the second element proves some property about the first element.
+In our case, the first element is the status code, and the second element is the proof that it is one of the pre-defined status codes in our API.
+
+We will now show how we would implement for \mintinline{idris}{GET}:
+\begin{minted}{idris}
+  (Universe req, Universe res) => Universe (Handler req res) where
+    el (GET req res statuses) =
+      (DPair Int (\n => Elem n statuses), IO (el responseType))
+\end{minted}
+So now instead of just returning the IO action, we also return the status code, paired with its proof:
+\begin{minted}{idris}
+api : Handler EmptyU EmptyU
+api = GET VOID [200,401,404]
+
+-- A proof that 401 is an acceptable status code
+the401IsInStatusCodes : Elem 401 [200, 401, 404]
+the401IsInStatusCodes = There Here
+
+handler : el api
+handler = (MkDPair 401 (the401IsInStatusCodes), putStrLn "hey")
+\end{minted}
+We can now modify the router such that it can use the status code to write it to the response:
+\begin{minted}{idris}
+  route (GET responseType statuses)
+        (MkDPair status proof, handler)
+        url requestBody =
+    Just $ do
+      putStrLn (show status)
+      map (toResponse responseType) handler
+\end{minted}
 \subsection{A proof of non-overlapping routes}
-\emph{TODO}
+One of the weaknesess of Servant is that it is possible to define routes that overlap. It would be great if we could give a compiler error in Servis if two routes overlap.
+
+In Idris, we can describe propositions using types.
+And then allows us to make proofs of these propositions by implementing a program of that type.
+Also, Idris comes with a proof searcher that can automatically search for a proof for a certain proposition.
+It will do a best effort to find a proof, and if it can not find a proof in time the compiler halts with an error.
+If it did not find a proof, it does not mean there is no possible proof though.
+The proof searcher is naive.
+It will only search 100 levels of recursion deep, and it will not do contradictory proofs.
+So in some cases, the user will have to provide a proof himself.
+
+We would like to automatically prove the proposition that for every path that we add to our API datatype, it does not overlap with any other path in the API datatype.
+This way, it is a compiler error if paths in the API overlap.
+We call this proposition \emph{disjointness}.
+
+Query parameters make comparing URLs clumsy because there can be duplicate query parameters and query parameters can appear in any order.
+Hence, for simplicity sake, we will omit query parameters when testing this proposition.
+We do this by defining the following filter function:
+
+\begin{minted}{idris}
+wqp : Path a b c d -> Path a b c d
+wqp (Outputs handler) = Outputs handler
+wqp ((QueryParam name type) :> y) = wqp y
+wqp (a :> y) = a :> wqp y
+\end{minted}
+
+Now we only have to be concerned with \mintinline{idris}{Const} and \mintinline{idris}{Capture}.
+Because \mintinline{idris}{Capture} signifies a variable substitution, it will always conflict with a \mintinline{idris}{Const}.
+E.g.,  \texttt{/users/hey} conflicts with \texttt{/users/:var} because \texttt{:var} could be substituted by \texttt{hey}.
+
+This leads us to defining the following datatype, which allows us to prove that a \mintinline{idris}{PathPart} is disjoint if  two \mintinline{idris}{Const} values have different names:
+\begin{minted}{idris}
+data DisjointPP : PathPart capture query -> PathPart capture query -> Type where
+  ConstD : IsNo (decEq str str') -> DisjointPP (Const str) (Const str')
+\end{minted}
+
+To make a decision whether or not the string are equal at compile time, we use the \mintinline{idris}{DecEq} interface.
+\begin{minted}{idris}
+interface DecEq t where
+  decEq : DecEq t => (a:t) -> (b:t) -> Dec (a = b)
+
+||| Decidability. A decidable property either holds or is a contradiction.
+data Dec : Type -> Type where
+  ||| The case where the property holds
+  ||| @ prf the proof
+  Yes : (prf : prop) -> Dec prop
+  ||| The case where the property holding would be a contradiction
+  ||| @ contra a demonstration that prop would be a contradiction
+  No  : (contra : prop -> Void) -> Dec prop
+\end{minted}
+We then create a proposition that can ony be proven if  a decision is made that a property does not hold:
+\begin{minted}{idris}
+data IsNo : Dec a -> Type where
+  ItIsNo : IsNo (No x)
+\end{minted}
+
+The fact that we can not construct a value of \mintinline{idris}{DisjointPP (Const x) (Capture x y)} indicates that a \mintinline{idris}{Const}, and a \mintinline{idris}{Capture} always overlap, which is the behaviour that we wanted to express.
+
+Showing that a \mintinline{idris}{Path} is disjoint has more cases.
+\begin{minted}{idris}
+data DisjointPath : Path capture query req res ->
+                    Path capture query req res ->
+                    Type where
+\end{minted}
+First of all, if two path parts are disjoint, a path that starts with these path parts is disjoint:
+\begin{minted}{idris}
+  PBasePP : DisjointPP pp1 pp2 -> DisjointPath (pp1 : p1) (pp2 : p2)
+\end{minted}
+Next, if a path that is not just an \mintinline{idris}{Outputs} is compared to a path that i just an \mintinline{idris}{Outputs}, then these paths are disjoint:
+\begin{minted}{idris}
+  PBaseOutputs1 : DisjointPath (pp :> p) (Outputs h)
+  PBaseOutputs2 : DisjointPath (Outputs h) (pp :> p)
+\end{minted}
+Now we can prove an entire path disjoint by induction:
+\begin{minted}{idris}
+  PStep : (inducHypth : DisjointPath p1 p2) -> DisjointPath (pp1 :> p1) (pp2 : p2)
+\end{minted}
+
+Now every time we add a path to an API, we want to prove that each path in API is disjoint from the path we want to add.
+Luckily, Idris comes with a proposition named \mintinline{idris}{All} that checks if a proposition holds for all elements in a list:
+\begin{minted}{idris}
+data All : (P : a -> Type) -> List a -> Type where
+  Nil : {P : a -> Type} -> All P Nil
+  (::) : {P : a -> Type} -> {xs : List a} -> P x -> All P xs -> All P (x :: xs)
+\end{minted}
+
+We can now define an operator that adds a path to an existing lists of paths, but only if we are able to provide a proof that the path is not in that list:
+\begin{minted}{idris}
+  (+>) : (path : Path c q rq rs) ->
+         (xs   : List (Path c q rq rs)) ->
+         {auto all : All (\x => DisjointPath (wqp path) (wqp x)) xs} ->
+         List (Path c q rq rs)
+  (+>) x xs = x::xs
+\end{minted}
+The \mintinline{idris}{auto} keyword signifies that idris will try to automatically find a value fo the implicit argument \mintinline{idris}{all}.
+If it fails, it will give a compiler error.
+But if it succeeds, we can be sure that we did not add an overlapping path to our list of paths.
+
+We can now make sure that our API definitions are sane. This will compile:
+\begin{minted}{idris}
+  api' : List (Path CapU EmptyU EmptyU RespU)
+  api' =  Const "users" :> Capture "name" STRING :> Outputs (GET USER)
+       +> Const "users" :> Capture "name" STRING
+                        :> Const "friends" :> Outputs (GET (LIST USER))
+       +> Nil
+\end{minted}
+
+However, if we introduce an overlap, the compiler will tell us that this
+api definition is invalid and that we need to fix our code:
+\begin{minted}{idris}
+apiOverlap : List (Path CapU EmptyU EmptyU RespU)
+apiOverlap
+  =  Const "users" :> Capture "name" STRING :> Outputs (GET USER)
+  +> Const "users" :> Const "all"           :> Outputs (GET (LIST USER))
+  +> Nil
+
+When checking right hand side of apiOverlap' with expected type
+        List (Path CapU EmptyU EmptyU RespU)
+
+When checking argument ok to function Servis.API.+>:
+  Can't find a value of type 
+    All (\x =>
+          DisjointPath (Const "users" :> Capture "name" STRING :> Outputs (GET USER))
+                      (wqp x))
+        [Const "users" :> Const "all" :> Outputs (GET USER)]
+\end{minted}
+
 \subsection{Adding API interdependencies}
@@
 And lets modify the domain function that queries the database for users
-in Listing~\ref{code:domain} to return bounded lists:
+in Listing~\ref{code:domain} to return a bounded list that is \emph{dependent} on the \mintinline{idris}{limit} parameter.
 
@@
 \end{minted}
-We can now describe an API in which we can describe the dependency between the query parameter
+We can now describe an API in which we can express the dependency between the query parameter
 and the size of the returned list of users.
@@
 will only show the implementation for \mintinline{idris}{QueryParam},
-but the changes to make \mintinline{idris}{Capture} work are identical.
+but the changes to make \mintinline{idris}{Capture} work are almost identical and left as an exercise to the reader.
 
@@
     -- BEFORE :
---  route right 
---        (handler param)
---        (MkURL pathParts newParams)
---         requestBody
-    -- WITH ADDED DEPENDENCY:
     route (right param) -- <- note the extra argument passed to RHS
@@
           requestBody
-\end{minted}
-And that's it. we now have added dependent types to our API!
-
-\subsection{New problems that dependent types create}
+
+\end{minted}
+
+\subsubsection{New problems that dependent types create}
 The dependent types are nice, but our datatype is now a Higher-order abstract syntax instead of a first-order abstract syntax.
@@
 The real problem though is that this representation breaks the overlapping path detector.
-The overlapping path detector uses the feature of idris to search proofs automatically. But that
-proof search isn't perfect, and won't be able to prove that paths overlap or not if we have
-arbitrary functions in our datatypes.
-
-A solution this problem would be that library authors would have to provide their own proof
-of whether or not their api definitions overlap, but this can be very cumbersome and thus
-is undesired.
-
+The overlapping path detector uses the feature of idris to search proofs automatically. But
+proof search isn't perfect will be exhausted after a while. Take ourexample with the
+\mintinline{idris}{limit : Nat} parameter. For Idris to find a proof, it would try every number
+in \mintinline{idris}{Nat} to  make sure that the part of our path after the dependency does
+not overlap. However, there are a countable infinite number of \mintinline{idris}{Nat}s so
+Idris would never be able to complete the proof. Hence the auto-searcher will stop searching
+for a proof after a while and quit.
+
+A solution is that library authors can provide the proof themselves. Maybe they can prove with
+additional information that the numbers never get higher than $8$, and that it thus is safe
+to try ever number from $0$ to $8$ to make sure no overlapping path can be generated. However
+writing such proofs can be cumbersome and is thus undesired.
 \section{Related work}
@@
 a closed-kinded Servant alternative. When I stumbled upon it, I got very curious
-what would happen if we would port it to Idris.  The serv source code is kind of
+what would happen if we would write something similar Idris.  The serv source code is kind of
 dense because Haskell isn't a very nice language to do type-level programming in

### 2016-08-25 paper/paper.tex
Final draft

@@
 We can accomplish this by letting the universe implementation return a dependent pair.
-A dependent pair in idris is a tuple where the second element proves some property about the first element.
+A dependent pair in Idris is a tuple where the second element proves some property about the first element.
 In our case, the first element is the status code, and the second element is the proof that it is one of the pre-defined status codes in our API.
@@
 \end{minted}
-The \mintinline{idris}{auto} keyword signifies that idris will try to automatically find a value fo the implicit argument \mintinline{idris}{all}.
+The \mintinline{idris}{auto} keyword signifies that Idris will try to automatically find a value fo the implicit argument \mintinline{idris}{all}.
 If it fails, it will give a compiler error.
@@
 A good usecase for these kind of dependent types is for endpoints that return a \mintinline{idris}{BoundedList}.
-A bounded list in idris is a list that can only contain upto a certain amount of elements. It is defined as follows:
+A bounded list in Idris is a list that can only contain upto a certain amount of elements. It  has the following signature:
 \begin{minted}{idris}
 data BoundedList : Nat -> Type -> Type where
-  Nil : BoundedList n a
-  (::) : a -> BoundedList n a -> BoundedList (S n) a
 \end{minted}
@@
 
-A solution is that library authors can provide the proof themselves. Maybe they can prove with
-additional information that the numbers never get higher than $8$, and that it thus is safe
-to try ever number from $0$ to $8$ to make sure no overlapping path can be generated. However
-writing such proofs can be cumbersome and is thus undesired.
+A solution is that library authors can provide the proof themselves, similarly to how this
+was solved for explicit status code checks.  However, writing these proofs for paths might
+defeat the point. We wanted them to be automatic to have a sanity check to make sure
+people are not caught off guard if certain paths are not routed. Having to write a proof
+for each dependent path you define makes the library cumbersome to use and thus might
+be undesired.
+
+\section{Conclusions}
+Idris allows us to elegently describe APIs as simple terms. We can
+then use the Universe pattern to calculate types for handler.
+These handler functions can then be made routable such that we have
+a function web server.
+
+Idris makes it easier to write utility functions, like documentation generation
+because our API definition does not live in the type-level as it did in Servant.
+
+We were able to solve several shortcomings of servant: By not using an open kind, 
+we eliminated the possibility of writing APIs that are not interpretable.  We also also the possibility
+of defining overlapping routes. W made status codes explicit in the API and gave
+library authors the tools to prove that their servers adhere to these status code
+definitions.  Finally, we were able to implement an dependent API, that allows interdependencies
+within paths to be able to describe even more powerful APIs.
+
 \section{Related work}
-This project was heavily inspired by \url{https://github.com/tel/serv}. Which is
+This project started with me looking at\url{https://github.com/tel/serv}. Which is
 a closed-kinded Servant alternative. When I stumbled upon it, I got very curious
@@
 dense because Haskell isn't a very nice language to do type-level programming in
-and it got me curious whether or not idris would improve readability. Which it did!
+and it got me curious whether or not Idris would improve readability. Which it did!
 
@@
 route conflict resolution.
-
-\section{Conclusions and future work}
-Idris allows us to elegantly describe APIs with plain ol' datatypes. We
-can then use the Universe pattern to map these datatypes to handler
-function types. These handler function types can then be made routable
-by writing an interpreter of the API datatype that decodes the incoming
-requests and passes the correct parameters to the handlers.
-
-Idris makes it easier to write utility functions, like documentation generation
-or code generation, because our API definition does not live in the type-level
-anymore, because type-level and term-level are united in idris.
-
-We were able to solve the common mistake of overlapping routes in Servant as well,
-by using Idris's proof search capabilities.
-
-Finally, we were able to implement an dependent API, that allows interdependencies
-within paths to be able to describe even more powerful APIs.
-
-In the future, it would be interesting to look at if it's possible to enforce in 
-the type-system that handlers only return HTTP Status codes that are defined
-in the API description. I explored some ideas on how to implement this, but
-did not have the time to do an implementation of this.
 \bibliography{sources}

### 2016-08-25 paper/paper.tex
Final final draft

@@
 
-TODO reword
 \begin{itemize}
   \item We point out several problems and limitations of the original Servant DSL
-  \item We show a port of servant to Idris, a dependently-typed language
-  \item We show how new techniques from Idris can help mitigate some of the problems that Servant has.
+  \item We show a port of Servant to Idris, a dependently-typed language
+  \item We show that Idris gives us new techniques to tackle these identified problems
 \end{itemize}
@@
 
-% TODO add text
 \begin{minted}{idris}
@@
 \end{minted}
-% TODO reword a bit
 
@@
 \end{minted}
-% TODO reword
-% TODO CHECKED UNTIL HERE. PAGE 11
-For example, we could use this to encode values in the \mintinline{idris}{ReqResUniverse} universe from
+We could use this to encode values in the \mintinline{idris}{ReqResUniverse} universe from
 Listing~\ref{code:api}:

### 2016-08-25 paper/paper.tex
Typos

@@
 Servis is  inspired by Servant, a haskell type-level DSL for web APIs.
-We identify several issues with the current implementation of Servantm and show how Servis solves these by using features of the dependently typed language called Idris.
-
-Specifically, we show how dependent types can help use describe dependencies within an API, how we can check at  compile time that routes in an API dont overlap, and how dependent pairs can let us verify that status codes described in an API are adhered to.
+We identify several issues with the current implementation of Servant and show how Servis solves these by using features of the dependently typed language called Idris.
+
+Specifically, we show how dependent types can help use describe dependencies within an API, how we can check at  compile time that routes in an API do not overlap, and how dependent pairs can let us verify that status codes described in an API are adhered to.
 

### 2016-08-25 paper/paper.tex
FINAL FINAL FINAL DRAFT

@@
 \author{Arian van Putten \\ a.vanputten@uu.nl}
-\title{Servis: A dependently typed DSL for web APIs}
+\title{Servis: A dependently typed DSL for web APIs\\
+\large Honours Thesis}
 \begin{document}
@@
 In this paper we describe a DSL for web APIs named Servis.
-Servis is  inspired by Servant, a haskell type-level DSL for web APIs.
+Servis is  inspired by Servant, a Haskell type-level DSL for web APIs.
 We identify several issues with the current implementation of Servant and show how Servis solves these by using features of the dependently typed language called Idris.
@@
 
-Though haskell has type-level functions (called type-families), it does not allow these functions to depend on term-level values.
+Though Haskell has type-level functions (called type-families), it does not allow these functions to depend on term-level values.
 There is a distinct boundary between type-level and term-level, whilst the two are merged in Idris.
@@
 \subsection{Contributions}
-In this paper, I make the following contributions:
+In this paper, we make the following contributions:
 
@@
 
+\subsection{Overlapping routes}
+Another problem with Servant is that there is no protection for overlapping routes.
+If two routes resolve to the same path, Servant makes a left-biased choice~\cite{servant}.
+It would be useful that instead of making a design choice where some routes are shadowed without warning, the compiler could inform the user about a potential mistake.
+
+\begin{listing}
+\begin{minted}{haskell}
+  type API  = "users" :> Capture "args" String :> Get [JSON] User
+         :<|> "users" :> "favorites" :> GET [JSON] User
+\end{minted}
+\caption{An example of overlapping routes. The capture will be chosen if args equals "favorites"}
+\end{listing}
+
+\subsection{Explicit status codes}
+It is great that Servant lets us give a presice description about the structure and types of our web APIs. However, it does now allow us to describe any form of errors within the API types.
+
+It is important for a developer to know how an API call could fail, such that he can handle it accordingly.
+Also, when generating documentation from an API type, it one would like to have this information at hand to inform the users of your API on what kind of errors they can expect.
+
+This would not be hard to add to Servant.
+A type-level list of error types paired with the return type of an API endpoint would suffice.
+However, it would also be interesting to explore if we can enforce that the code that implements a server for an API actually only returns the error codes that are documented.
+
 \subsection{API Interdependencies}
@@
 
-\subsection{Overlapping routes}
-Another problem with Servant is that there is no protection for overlapping routes.
-If two routes resolve to the same path, Servant makes a left-biased choice~\cite{servant}.
-It would be useful that instead of making a design choice where some routes are shadowed without warning, the compiler could inform the user about a potential mistake.
-
-\begin{listing}
-\begin{minted}{haskell}
-  type API  = "users" :> Capture "args" String :> Get [JSON] User
-         :<|> "users" :> "favorites" :> GET [JSON] User
-\end{minted}
-\caption{An example of overlapping routes. The capture will be chosen if args equals "favorites"}
-\end{listing}
-
-\subsection{Explicit status codes}
-It is great that Servant lets us give a presice description about the structure and types of our web APIs. However, it does now allow us to describe any form of errors within the API types.
-It is important for a developer to know how an API call could fail, such that he can handle it sufficently.
-Also, when generating documentation from an API type, it one would like to have this information at hand to inform the users of your API on what kind of errors they can expect.
-
-This would not be hard to add to Servant.
-A type-level list of error types paired with the return type of an API endpoint would suffice.
-However, it would also be interesting to explore if we can enforce that the code that implements a server for an API, actually only returns the error codes that are documented.
-
 \section{Servis}
@@
 It tries to address each issue listed in Section~\ref{sec:issues}.
-Firstly, we will define the DSL in which we can write API specifications. Then, we will look at how to write an interpreter of this DSL that generates documentation.
-Then, we will look at how we would implement HTTP webserver handlers that adhere to specification defined by the API DSL\@.
-Next, we'll look at solving the overlapping routes problem by using Idris's automatic proof search.
+Firstly, we will define the DSL in which we can write API specifications in a closed kind so that no malformed
+descriptions can exist. Then, we will look at how to write an interpreter of this DSL that generates documentation.
+Next, we will look at solving the overlapping routes problem by using Idris's automatic proof search.
+Then, we will  extend the DSL to define accepted status coded, and explore a way to check if handlers adhere to this definition.
 Finally, we will extend the DSL to add dependent types, allowing us to describe dependencies in our API\@.
-For example, one might want to describe  the size of a list of users in the API, but let the size depend on a query parameter \emph{limit} that limits the response size.
-It would then force the implementor of the API to only return values smaller or equal to limit in size.
 
@@
 
-Sadly enough, this code will give us a compiler error. The problem is that in Idirs, we cannot pattern match on values of \mintinline{idris}{Type}. 
+Sadly enough, this code will give us a compiler error. The problem is that in Idris, we cannot pattern match on values of \mintinline{idris}{Type}. 
 However, this exact problem can be solved with universes We can define a data type \mintinline{idris}{ResponseUniverse : Type} which names
@@
 Of course we do not only get request information from the request body,
-we also get information about a request by parsing it's request URL. This
+we also get information about a request by parsing its request URL. This
 is where \mintinline{idris}{Path}s and \mintinline{idris}{PathPart}s come into play. 
@@
 want to describe the names of the parts of these paths, but also their types. Hence, we will parameterize
-over a universe similarly to how we defined \mintinline{idris}{Handler}.
+over a universe similarly to how we did in the definition of \mintinline{idris}{Handler}.
 \begin{minted}{idris}
@@
 Writing an interpreter for generating documentation is straightforward.
-Our API description isn't some type-level definition, it's just a datatype that can be passed to functions and can be manipulated.
-
-We start by introducing an interface for docs, which just tells us how to turn something into a string:
+Our API description is not just some type-level definition, it is  a datatype that can be passed to functions and can be manipulated.
+
+We start by introducing an interface for docs, which simply tells us how to turn something into a string:
 \begin{minted}{idris}
@@
 \end{minted}
-Now, we just need to define an implementation for \mintinline{idris}{API} for this interface and we are done.
+Now, we need to define an implementation for \mintinline{idris}{API} for this interface and we are done.
 To implement this, we will of course also needs to implement the interface for \mintinline{idris}{Handler}, \mintinline{idris}{PathPart} and \mintinline{idris}{Path}.
 
-To get an idea on how this would work, we will show the implementation of \mintinline{idris}{Handler} and \mintinline{idris}{API}:
+To get an idea on how this would loook like, we will show the implementation of \mintinline{idris}{Handler} and \mintinline{idris}{API}. The others are left as an exercise for the reader:
 \begin{minted}{idris}
@@
 
-We will more-or-less use this representation to model our HTTP handlers.An actual implementation of a web server based on this model is out of scope for now, but it would be interesting to port something like Haskells' \texttt{warp} (which impelements this concept) to Idris.
+We will more-or-less use this representation to model our HTTP handlers.
+An actual implementation of a web server based on this model is out of scope for now, but it would be interesting to port something like Haskell's \texttt{warp} web server (which impelements this concept) to Idris.
 
@@
 match the API path with the incoming request. In case of a mismatch, the router should then try if the next available path does match the request.
-Instead of letting the application return a value of \mintinline{idris}{IO String}, let it return a value of \mintinline{idris}{Maybe (IO String)}, where \mintinline{idris}{Just a} signals a route match, whilst \mintinline{idris}{Nothing} signals a route mistmatch such that we can try another route.
+Instead of letting the application return a value of \mintinline{idris}{IO String}, we let it return a value of \mintinline{idris}{Maybe (IO String)}, where \mintinline{idris}{Just a} signals a route match, and \mintinline{idris}{Nothing} signals a route mistmatch such that we can try another route.
 
@@
 The first part of the API for which we will implement the routing mechanics is the \mintinline{idris}{Handler}.
-Defining this router consists of two cases: A \mintintline{idris}{GET} or a \mintinline{idris}{POST}.
+Defining this router consists of two cases: A \mintinline{idris}{GET} or a \mintinline{idris}{POST}.
 During a \mintinline{idris}{GET}, we just have to execute the handler which will then give us a response. 
@@
 Routing multiple paths is easy. We just keep on trying consecutive paths until one
-does nott return \mintinline{idris}{Nothing}:
+does not return \mintinline{idris}{Nothing}:
 \begin{minted}{idris}
@@
 
-\subsection{Explicit status codes}
-To better document the behaviour of our API to users of the API, we can add a list of status codes that an endpoint can return by modifing \mintinline{idris}{Handler}:
-
-\begin{minted}{idris}
-data Handler : (req : Type)  -> (res : Type) -> Type where
-  GET : (responseType : res) -> (statuses : List Int) -> Handler req res
-  POST : (requestType : req) -> (responseType : res) ->
-         (statuses : List Int)  -> Handler req res
-\end{minted}
-
-However, it would be convenient if we could automatically deduce a proof that our handlers never return
-other status codes than are in that list. However, this gets impossible quickly because not all programs are decidable.
-Instead, we can let our \mintinline{idris}{Universe} implementation generate a type that forces the implementor of the API to provide a proof for us.
-We can accomplish this by letting the universe implementation return a dependent pair.
-A dependent pair in Idris is a tuple where the second element proves some property about the first element.
-In our case, the first element is the status code, and the second element is the proof that it is one of the pre-defined status codes in our API.
-
-We will now show how we would implement for \mintinline{idris}{GET}:
-\begin{minted}{idris}
-  (Universe req, Universe res) => Universe (Handler req res) where
-    el (GET req res statuses) =
-      (DPair Int (\n => Elem n statuses), IO (el responseType))
-\end{minted}
-So now instead of just returning the IO action, we also return the status code, paired with its proof:
-\begin{minted}{idris}
-api : Handler EmptyU EmptyU
-api = GET VOID [200,401,404]
-
--- A proof that 401 is an acceptable status code
-the401IsInStatusCodes : Elem 401 [200, 401, 404]
-the401IsInStatusCodes = There Here
-
-handler : el api
-handler = (MkDPair 401 (the401IsInStatusCodes), putStrLn "hey")
-\end{minted}
-We can now modify the router such that it can use the status code to write it to the response:
-\begin{minted}{idris}
-  route (GET responseType statuses)
-        (MkDPair status proof, handler)
-        url requestBody =
-    Just $ do
-      putStrLn (show status)
-      map (toResponse responseType) handler
-\end{minted}
 \subsection{A proof of non-overlapping routes}
-One of the weaknesess of Servant is that it is possible to define routes that overlap. It would be great if we could give a compiler error in Servis if two routes overlap.
+One of the weaknesess of Servant is that it is possible to define routes that overlap.
+It would be great if we could give a compiler error in Servis if this is the case.
 
 In Idris, we can describe propositions using types.
-And then allows us to make proofs of these propositions by implementing a program of that type.
-Also, Idris comes with a proof searcher that can automatically search for a proof for a certain proposition.
+This then allows us to make proofs of these propositions by implementing a program of that type.
+Idris also comes with a proof searcher that can automatically search for a proof for a certain proposition.
 It will do a best effort to find a proof, and if it can not find a proof in time the compiler halts with an error.
@@
 E.g.,  \texttt{/users/hey} conflicts with \texttt{/users/:var} because \texttt{:var} could be substituted by \texttt{hey}.
+It will also conflict with another \mintinline{idris}{Capture}.
 
@@
   ConstD : IsNo (decEq str str') -> DisjointPP (Const str) (Const str')
+
 \end{minted}
@@
 
-The fact that we can not construct a value of \mintinline{idris}{DisjointPP (Const x) (Capture x y)} indicates that a \mintinline{idris}{Const}, and a \mintinline{idris}{Capture} always overlap, which is the behaviour that we wanted to express.
+The fact that we can not construct a value of \mintinline{idris}{DisjointPP (Const x) (Capture x y)} nor a \mintinline{idris}{DisjointPP (Capture x y) (Capture z f)} indicates that these will always overlap, which is the behaviour that we wanted to express.
 
@@
 However, if we introduce an overlap, the compiler will tell us that this
-api definition is invalid and that we need to fix our code:
+API definition is invalid and that we need to fix our code:
 \begin{minted}{idris}
@@
         [Const "users" :> Const "all" :> Outputs (GET USER)]
+\end{minted}
+
+\subsection{Explicit status codes}
+To better document the behaviour of our API to users of the API, we can add a list of status codes that an endpoint can return by modifing \mintinline{idris}{Handler}:
+
+\begin{minted}{idris}
+data Handler : (req : Type)  -> (res : Type) -> Type where
+  GET : (responseType : res) -> (statuses : List Int) -> Handler req res
+  POST : (requestType : req) -> (responseType : res) ->
+         (statuses : List Int)  -> Handler req res
+\end{minted}
+
+However, it would be convenient if we could automatically deduce a proof that our handlers never return
+other status codes than are in that list. However, this gets impossible quickly because not all programs are decidable.
+Instead, we can let our \mintinline{idris}{Universe} implementation generate a type that forces the implementor of the API to provide a proof for us.
+We can accomplish this by letting the universe implementation return a dependent pair.
+A dependent pair in Idris is a tuple where the second element proves some property about the first element.
+In our case, the first element is the status code, and the second element is the proof that it is one of the pre-defined status codes in our API.
+
+We will now show how we would implement this for \mintinline{idris}{GET}:
+\begin{minted}{idris}
+  (Universe req, Universe res) => Universe (Handler req res) where
+    el (GET req res statuses) =
+      (DPair Int (\n => Elem n statuses), IO (el responseType))
+\end{minted}
+So now instead of just returning the IO action, we also return the status code, paired with its proof:
+\begin{minted}{idris}
+api : Handler EmptyU EmptyU
+api = GET VOID [200,401,404]
+
+-- A proof that 401 is an acceptable status code
+the401IsInStatusCodes : Elem 401 [200, 401, 404]
+the401IsInStatusCodes = There Here
+
+handler : el api
+handler = (MkDPair 401 (the401IsInStatusCodes), putStrLn "hey")
+\end{minted}
+We can now modify the router such that it can use the status code to write it to the response:
+\begin{minted}{idris}
+  route (GET responseType statuses)
+        (MkDPair status proof, handler)
+        url requestBody =
+    Just $ do
+      putStrLn (show status)
+      map (toResponse responseType) handler
 \end{minted}
@@
 \subsubsection{New problems that dependent types create}
-The dependent types are nice, but our datatype is now a Higher-order abstract syntax instead of a first-order abstract syntax.
-Which means the datatype contains a function.  This makes it difficult to generate documentation as we need to provide a value
-to apply to the function to traverse the datatype. It isn't impossible though. We could introduce the following interface:
+The dependent types are nice, but now our datatype contains a function.
+This makes it difficult to generate documentation as we need to provide a value
+to apply to the function to traverse the datatype. It is not impossible though. We could introduce the following interface:
 \begin{minted}{idris}
@@
 The overlapping path detector uses the feature of idris to search proofs automatically. But
-proof search isn't perfect will be exhausted after a while. Take ourexample with the
+proof search is not perfect and will be exhausted after a while. Take our example with the
 \mintinline{idris}{limit : Nat} parameter. For Idris to find a proof, it would try every number
@@
 definitions.  Finally, we were able to implement an dependent API, that allows interdependencies
-within paths to be able to describe even more powerful APIs.
-
-\section{Related work}
-This project started with me looking at\url{https://github.com/tel/serv}. Which is
+within paths to be able to describe even more powerful APIs. However, these dependent types do make
+it more difficult to write implementations for APIs.  A users will have to provide proofs that his
+handlers adhere to the pre-defined status codes, and will also have to prove manually that his
+routes do not overlap.
+
+\section{Related work and Future work}
+This project started with me looking at \url{https://github.com/tel/serv}. Which is
 a closed-kinded Servant alternative. When I stumbled upon it, I got very curious
-what would happen if we would write something similar Idris.  The serv source code is kind of
-dense because Haskell isn't a very nice language to do type-level programming in
-and it got me curious whether or not Idris would improve readability. Which it did!
-
-Also, once I had a working example of Serv in idris, I wanted to see if idris
-could add any new tricks to it. This made me implement dependent types and
-route conflict resolution.
+what would happen if we would write something similar in a dependently typed language.
+I had never used a dependently typed language before, or really thought of what it would gain me.
+I decided to learn Idris, and discovered that it would make implementing a Servant alternative
+very elegant.
+
+When I got an implementation working, I got curious if I could add more exotic features using Idris.
+The question now is whether the things I have discovered could be ported back to serv or Servant 
+once dependent types get added to GHC. This would be great to explore in the future.
 \bibliography{sources}
